---
title: "Wearable Behavioral Phenotyping in Adolescence"
subtitle: "Fitbit-derived activity and sleep phenotypes and their associations with mental-health and metabolic outcomes (ABCD Study). Detailed Methods, Results, and Discussion."
author: "Biplabendu Das"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    fig-width: 8
    fig-height: 5
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
  freeze: auto
---

```{r setup}
library(targets)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
library(scales)
knitr::opts_chunk$set(dpi = 150)

# --- Materialized pipeline outputs ------------------------------------------
cfg              <- tar_read(cfg)
master           <- tar_read(master)
qc_compliance    <- tar_read(qc_compliance)
alignment        <- tar_read(alignment)
transitions      <- tar_read(transitions)
gbtm             <- tar_read(gbtm)
mh_feature_summary <- tar_read(mh_feature_summary)
mh_traj_model    <- tar_read(mh_traj_model)
mh_models        <- tar_read(mh_models)
metabolic_models <- tar_read(metabolic_models)
multivariate     <- tar_read(multivariate)
fig_umap         <- tar_read(fig_umap)
fig_diurnal      <- tar_read(fig_diurnal)
fig_alluvial     <- tar_read(fig_alluvial)
fig_canonical    <- tar_read(fig_canonical)

# --- Formatting helpers -----------------------------------------------------
msd <- function(x, d = 2) {
  sprintf(paste0("%.", d, "f (%.", d, "f)"), mean(x, na.rm = TRUE), stats::sd(x, na.rm = TRUE))
}
fmt_p  <- function(p) if (is.na(p)) "NA" else if (p < 0.001) "p < 0.001" else paste0("p = ", signif(p, 2))
fmt_pf <- function(p) if (is.na(p)) "NA" else if (p < 0.001) "< 0.001" else signif(p, 2)
gv <- function(tbl, term, col) { i <- which(tbl$term == term); if (!length(i)) NA_real_ else tbl[[col]][i[1]] }
fs_get <- function(f, col) { r <- mh_feature_summary[mh_feature_summary$feature == f, ]; if (nrow(r)) r[[col]] else NA_real_ }

# --- Headline quantities ----------------------------------------------------
n_pw    <- nrow(master)
n_part  <- dplyr::n_distinct(master$participant_id)
n_clust <- nlevels(master$cluster_aligned)
waves   <- paste(sort(unique(master$wave_label)), collapse = " and ")
signals <- paste(names(cfg$signals), collapse = ", ")

al_min <- min(alignment$cosine_similarity)
al_max <- max(alignment$cosine_similarity)
al_low <- alignment$aligned_cluster[which.min(alignment$cosine_similarity)]

# primary feature × age effects
on_b  <- fs_get("z_min_slp_onset_sd", "beta_x_age");      on_fdr <- fs_get("z_min_slp_onset_sd", "p_x_age_fdr")
st_b  <- fs_get("z_steps_total_mean_daily", "beta_x_age"); st_fdr <- fs_get("z_steps_total_mean_daily", "p_x_age_fdr")
mv_b  <- fs_get("z_steps_total_mvpa_fraction", "beta_x_age"); mv_fdr <- fs_get("z_steps_total_mvpa_fraction", "p_x_age_fdr")

# trajectory & wave-nested cluster
trw_b <- gv(mh_traj_model$table, "prob_worsening:wave_age", "Estimate")
trw_p <- gv(mh_traj_model$table, "prob_worsening:wave_age", "Pr(>|t|)")

# Dynamic: find the yr6 cluster term with the smallest p-value (most significant)
yr6_terms  <- mh_models$burden |>
  dplyr::filter(grepl("^wave_labelyr6:cluster_aligned", term))
yr6_sig    <- if (nrow(yr6_terms) > 0) yr6_terms$term[which.min(yr6_terms[["Pr(>|t|)"]])] else NA_character_
yr6_cl_id  <- if (!is.na(yr6_sig)) sub("wave_labelyr6:cluster_aligned", "", yr6_sig) else "?"
yr2_term   <- sub("yr6", "yr2", yr6_sig)
cl6_b      <- if (!is.na(yr6_sig)) gv(mh_models$burden, yr6_sig,  "Estimate")    else NA_real_
cl6_p      <- if (!is.na(yr6_sig)) gv(mh_models$burden, yr6_sig,  "Pr(>|t|)")   else NA_real_
cl2_p      <- if (!is.na(yr6_sig)) gv(mh_models$burden, yr2_term, "Pr(>|t|)")   else NA_real_

# Dynamic: count FDR-significant primary features
n_sig_feat   <- sum(
  c(on_fdr, st_fdr, mv_fdr,
    fs_get("z_min_slp_duration_mean", "p_x_age_fdr")) < 0.05, na.rm = TRUE
)
# Dynamic: count well-preserved phenotypes (all except the minimum-similarity one)
n_preserved  <- sum(alignment$cosine_similarity > al_min)

# transitions & multivariate
tr_p  <- transitions$chisq$p.value
cc1   <- multivariate$canonical_cor[1]
man_p <- tryCatch(multivariate$manova$stats[1, "Pr(>F)"], error = function(e) NA_real_)

# --- Unclustered / noise participant counts ----------------------------------
n_noise_pv  <- sum(is.na(master$cluster_aligned))
n_clust_pv  <- sum(!is.na(master$cluster_aligned))
n_noise_ids <- master |>
  dplyr::filter(is.na(cluster_aligned)) |>
  dplyr::distinct(participant_id) |>
  nrow()

# --- Wear-quality CSV for attrition table -----------------------------------
wq_path <- file.path(cfg$handoff_dir, "wear_quality.csv")
wq      <- tryCatch(readr::read_csv(wq_path, show_col_types = FALSE), error = function(e) NULL)
n_wq_pv  <- if (!is.null(wq)) nrow(wq)                             else NA_integer_
n_wq_ids <- if (!is.null(wq)) dplyr::n_distinct(wq$participant_id) else NA_integer_
```

# Methods

## Sample

Participants were adolescents in the Adolescent Brain Cognitive Development (ABCD)
cohort with concurrent Fitbit sleep and activity data, restricted to a curated set
with usable sleep and step data at both analysed visits. Devices recorded for a
21-day window per visit; records were pre-aggregated to **2-hour epochs** (12
slots/day). Two visits are analysed — **ses-02A (yr2, ~age 12)** and **ses-06A
(yr6, ~age 16)**. The analytic file contains
`r format(n_pw, big.mark = ",")` participant-visits from
**`r n_part` participants**.

## Wearable signals

Three complementary 2-hour-resolution signals were used: **`r signals`** —
i.e. step volume (`steps_total`), Metabolic Equivalents of Task (METs; `mets`), and
minutes of sleep per slot (`min_slp`). Resting heart rate was excluded as
unreliable for this device/cohort; METs and intensity categories, though
heart-rate–derived for quality control, are valid activity measures and were
retained.

## Preprocessing

A four-stage Python pipeline, run once and logged: (1) **artifact removal** —
per-signal physiological bounds and within-participant-day spike removal
(> mean + 5 standard deviations [SD]); (2) **within-day imputation** of gaps
≤ 3 consecutive slots by the same-time-of-day median over ± 3 days (≥ 3 valid
neighbours required); (3) **day-validity classification** — a slot counts as worn
when recorded minutes (`min_total`) > 0, and a day is valid with ≥ 8 of 12 worn
slots including ≥ 1 night-time slot; participants with fewer than 14 valid days
were excluded; (4) **feature extraction** from valid days only. Participants with
any missing z-scored feature (arising from sparse data in specific time slots) were
excluded from dimensionality reduction and clustering but retained in the
feature-level mixed models.

```{r tbl-attrition}
tibble::tibble(
  Stage = c(
    "Passed wear-quality screen (≥14 valid days at ≥1 visit)",
    "In analytic dataset (features joined to outcomes)",
    "  — assigned a cluster label",
    "  — missing features (excluded from clustering)"
  ),
  `N participants` = c(
    n_wq_ids, n_part,
    n_part - n_noise_ids,
    n_noise_ids
  ),
  `N participant-visits` = c(
    n_wq_pv, n_pw,
    n_clust_pv,
    n_noise_pv
  )
) |>
  kable(caption = "Table P1. Preprocessing attrition summary. Participants excluded from clustering (missing features) are retained in Tables 7–8 (feature-level mixed models) but absent from Tables 3, 9, and all cluster figures.",
        format.args = list(big.mark = ","))
```

## Feature extraction

For activity signals we computed:

- **Circadian rhythm indices** — single-harmonic cosinor amplitude and acrophase
  (peak time); Interdaily Stability (IS; between-day consistency of the 24-h
  rhythm pattern, 0–1); Intradaily Variability (IV; within-day fragmentation);
  L5 (mean activity in the least-active 5-hour window, 3 consecutive 2-h slots)
  and M10 (most-active 10-hour window, 5 consecutive 2-h slots); and relative
  amplitude (RA = [M10 − L5] / [M10 + L5]).
- **Daily summaries** — mean, SD, coefficient of variation (CV), and lag-1
  autocorrelation of daily totals; sedentary fraction and Moderate-to-Vigorous
  Physical Activity (MVPA) fraction of daytime slots.

For sleep signals: nightly duration (mean and SD); onset timing (circular mean
and SD, where onset = first slot with min_slp > 0 in the evening-to-morning
window); sleep efficiency (fraction of nighttime slots with detected sleep);
Social Jet Lag (SJL; free-night midsleep minus work-night midsleep, conventional
positive sign = later sleep on free nights); and day-to-day autocorrelation.

**Circadian profile construction for cluster characterization.** For each
participant-visit, the mean diurnal profile was computed by averaging each
2-hour slot value across all valid days (slot 0 = 00:00–02:00, …, slot 22 =
22:00–00:00). The cluster-level mean profiles shown in Figure 2 are the grand
mean of these per-participant profiles within each cluster, with 95% confidence
ribbons derived as ±1.96 × SE, where SE is the SD of participant-level slot means
divided by the square root of the number of participants in that cluster–visit
cell. Cosinor models were fitted to the cluster-mean profile to extract amplitude
and acrophase; IS and IV were computed from the day × slot matrix of individual
participants. All circadian indices are 2-hour-resolution approximations (see
Limitations).

Each feature was z-scored across participants within visit (population mean and SD
computed within that visit only, preventing cross-visit leakage); both raw and
z-scored versions were retained.

## Clustering and cross-wave alignment

Each visit was clustered independently using the pipeline:

1. **Principal Component Analysis (PCA)** — retaining components that explain 95%
   of variance in the z-scored feature matrix, reducing dimensionality while
   preserving most information.
2. **Uniform Manifold Approximation and Projection (UMAP)** — 10-component
   embedding used for clustering; separate 2-component embedding for visualization
   only (Figure 1).
3. **Hierarchical Density-Based Spatial Clustering of Applications with Noise
   (HDBSCAN)** — primary clustering; a parameter sweep over minimum cluster sizes
   and minimum samples selected the configuration with the highest silhouette score
   on non-noise points. Noise-labeled points (HDBSCAN label = −1) were resolved by
   assigning each to its nearest non-noise neighbor in UMAP-10 space (1-nearest
   neighbor). Participants remaining unclustered due to entirely missing feature
   vectors (upstream NaN; see Table P1) were excluded from this stage.
4. **k-medoids** (k = 3–7, silhouette-selected) — retained as a diagnostic
   comparison; its labels are not used in any reported analysis.
5. **Gaussian Mixture Model (GMM)** — fitted with k equal to the number of HDBSCAN
   clusters to provide soft cluster-membership probabilities per participant.

Within-wave stability was assessed by bootstrap resampling (100 iterations, 80%
subsample): the Adjusted Rand Index (ARI) between the full-data labels and each
resample's k-medoids solution at the same k was logged. Clusters were aligned
across visits by matching feature-space centroids (Hungarian algorithm on cosine
distance, reference = yr2); the **matched-pair cosine similarity is reported as a
diagnostic** of whether a label denotes the same phenotype at both ages.

## Outcomes and covariates

Outcome and covariate data were assembled with `NBDCtools`. The primary
mental-health outcome is **current-diagnosis burden** (count of present Kiddie
Schedule for Affective Disorders and Schizophrenia [KSADS] diagnoses); a binary
*any-depression* indicator is secondary. Metabolic outcomes are blood-panel labs
(`r paste(cfg$metabolic_outcome_vars, collapse = ", ")`).
Covariates are sex, age, and **pubertal stage** — the last formed by coalescing the
two sex-specific Pubertal Development Scale (PDS)-derived approximate-Tanner scores
into one ordinal scale.

## Statistical analysis

All outcome models adjust for sex, pubertal stage, and **wear-time fraction**
(a compliance covariate; see §3.1), with a participant random intercept for repeated
measures. The modelling strategy was chosen to respect the cross-wave instability of
cluster labels (§3.2):

- **Primary — behavioral feature × age (Linear Mixed Models [LMM]).**
  Because continuous features have a consistent meaning at every visit, four
  pre-specified features are each tested independently:

  $$\text{lmer}(\texttt{burden} \sim z\_\text{feature} \times \texttt{wave\_age} + \texttt{sex} + \texttt{pubertal\_stage} + \texttt{wear\_fraction} + (1 \mid \texttt{id}))$$

  The `z_feature:wave_age` interaction term tests developmental moderation
  (change in the feature–burden association per year of age). Benjamini–Hochberg
  (BH) False Discovery Rate (FDR) correction is applied across the four interaction
  p-values.

- **Secondary — trajectory group (LMM).** Group-Based Trajectory Modeling (GBTM)
  via `lcmm::hlme` on sleep-onset variability yields per-participant soft
  memberships; their posterior probabilities (reference = regular group) enter:

  $$\text{lmer}(\texttt{burden} \sim \texttt{prob\_worsening} \times \texttt{wave\_age} + \texttt{prob\_irregular} \times \texttt{wave\_age} + \texttt{covariates} + (1 \mid \texttt{id}))$$

  The Bayesian Information Criterion (BIC) selected the number of trajectory groups.

- **Descriptive — within-wave clusters (LMM).** Cluster effects nested within
  visit; no cross-wave label equivalence assumed:

  $$\text{lmer}(\texttt{burden} \sim \texttt{wave\_label} + \texttt{wave\_label}{:}\texttt{cluster\_aligned} + \texttt{covariates} + (1 \mid \texttt{id}))$$

  A parallel Generalized Linear Mixed Model (GLMM) with logistic link tests the
  binary *any-depression* outcome.

- **Multivariate.** Canonical Correlation Analysis (CCA) via `stats::cancor`
  between the z-scored feature matrix and the outcome set; Multivariate Analysis
  of Variance (MANOVA) tests overall cluster separation across features.

## Reproducibility

The Python environment is pinned (`conda`); R packages are locked with `renv`. The R
analysis is a `targets` DAG (`_targets.R`); every table and figure below is read
directly from the pipeline store, so the document regenerates exactly from the cached
results. A full clean rebuild (`targets::tar_make()`) requires the private ABCD Fitbit
data accessed via `NBDCtools` (`dir_abcd = "dev/data/fitbit-summaries/..."`), which is
not publicly distributed. External reproduction is limited to re-rendering from the
cached `_targets/` store; all cached objects were complete and current at analysis time.

# Results

## Exploratory data characteristics

Before modelling, we inspect the raw distributions of primary predictors and
outcomes to confirm data quality and describe the analytic sample.

```{r fig-pred-dist}
#| fig-cap: "Figure E1. Distribution of the four primary z-scored behavioral predictors by visit (yr2 ≈ age 12, yr6 ≈ age 16). Each panel shows a density curve per visit; the x-axis is the within-visit z-score (0 = visit mean, ±1 = one SD). Distributions are approximately normal after z-scoring; any residual skew is handled by the mixed-model's robustness to non-normality."
#| fig-height: 6
feat_labels <- c(
  z_min_slp_onset_sd          = "Sleep-onset variability (SD, z)",
  z_steps_total_mean_daily    = "Daily steps (z)",
  z_steps_total_mvpa_fraction = "MVPA fraction (z)",
  z_min_slp_duration_mean     = "Sleep duration (z)"
)
master |>
  dplyr::select(wave_label, dplyr::all_of(names(feat_labels))) |>
  tidyr::pivot_longer(-wave_label, names_to = "feature", values_to = "z") |>
  dplyr::filter(!is.na(z)) |>
  dplyr::mutate(feature = feat_labels[feature],
                feature = factor(feature, levels = feat_labels)) |>
  ggplot(aes(x = z, colour = wave_label, fill = wave_label)) +
  geom_density(alpha = 0.25, linewidth = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  facet_wrap(~feature, scales = "free_y", ncol = 2) +
  scale_colour_manual(values = c(yr2 = "#3B9AB2", yr6 = "#E07B39"), name = "Visit") +
  scale_fill_manual(  values = c(yr2 = "#3B9AB2", yr6 = "#E07B39"), name = "Visit") +
  labs(x = "Within-visit z-score", y = "Density") +
  theme_linedraw(12)
```

```{r fig-raw-pred-dist}
#| fig-cap: "Figure E2. Raw (unscaled) distributions of the four primary predictors by visit, showing the natural units before z-scoring. Vertical dotted lines mark visit medians."
#| fig-height: 6
raw_labels <- c(
  min_slp_onset_sd          = "Sleep-onset variability (SD of onset hour, h)",
  steps_total_mean_daily    = "Daily steps (count)",
  steps_total_mvpa_fraction = "MVPA fraction (proportion of daytime slots)",
  min_slp_duration_mean     = "Sleep duration (h/night)"
)
master |>
  dplyr::select(wave_label, dplyr::all_of(names(raw_labels))) |>
  tidyr::pivot_longer(-wave_label, names_to = "feature", values_to = "value") |>
  dplyr::filter(!is.na(value)) |>
  dplyr::mutate(feature = raw_labels[feature],
                feature = factor(feature, levels = raw_labels)) |>
  ggplot(aes(x = value, colour = wave_label, fill = wave_label)) +
  geom_density(alpha = 0.25, linewidth = 0.8) +
  stat_summary(aes(xintercept = after_stat(x), y = 0), fun = median,
               geom = "vline", linetype = "dotted", linewidth = 0.7) +
  facet_wrap(~feature, scales = "free", ncol = 2) +
  scale_colour_manual(values = c(yr2 = "#3B9AB2", yr6 = "#E07B39"), name = "Visit") +
  scale_fill_manual(  values = c(yr2 = "#3B9AB2", yr6 = "#E07B39"), name = "Visit") +
  labs(x = NULL, y = "Density") +
  theme_linedraw(12)
```

```{r fig-outcome-dist}
#| fig-cap: "Figure E3. Distribution of the primary (diagnosis burden count) and secondary (any-depression binary) mental-health outcomes by visit. Left: diagnosis burden is right-skewed (typical for count outcomes) and increases from yr2 to yr6. Right: prevalence of any depressive diagnosis by visit; bars show proportions with 95% Wilson confidence intervals."
#| fig-height: 4
library(patchwork)

p_burden <- master |>
  dplyr::filter(!is.na(mh_burden_present)) |>
  ggplot(aes(x = mh_burden_present, fill = wave_label)) +
  geom_bar(position = "dodge", alpha = 0.85, colour = "white") +
  scale_fill_manual(values = c(yr2 = "#3B9AB2", yr6 = "#E07B39"), name = "Visit") +
  scale_x_continuous(breaks = 0:10) +
  labs(x = "Diagnosis burden (count)", y = "N participant-visits",
       title = "Primary outcome") +
  theme_linedraw(11)

p_dep <- master |>
  dplyr::filter(!is.na(mh_any_depression)) |>
  dplyr::group_by(wave_label) |>
  dplyr::summarise(n = dplyr::n(),
                   k = sum(mh_any_depression == 1, na.rm = TRUE),
                   prop = k / n,
                   lo = prop.test(k, n)$conf.int[1],
                   hi = prop.test(k, n)$conf.int[2],
                   .groups = "drop") |>
  ggplot(aes(x = wave_label, y = prop, fill = wave_label)) +
  geom_col(alpha = 0.85, colour = "white", width = 0.5) +
  geom_errorbar(aes(ymin = lo, ymax = hi), width = 0.15) +
  scale_fill_manual(values = c(yr2 = "#3B9AB2", yr6 = "#E07B39"), guide = "none") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(x = "Visit", y = "Prevalence", title = "Secondary outcome") +
  theme_linedraw(11)

p_burden + p_dep + plot_layout(widths = c(2, 1))
```

## Sample characteristics and wear compliance

Participant characteristics by phenotype and visit are shown in Table 1. Clusters are
comparable in age and pubertal stage within each visit, indicating the phenotypes are
not simply age or puberty artefacts.

```{r tbl-participants}
master |>
  group_by(Visit = wave_label, Cluster = cluster_aligned) |>
  summarise(
    `N (visits)`     = dplyr::n(),
    `Female n (%)`   = sprintf("%d (%.0f)", sum(sex == "Female", na.rm = TRUE),
                               100 * mean(sex == "Female", na.rm = TRUE)),
    `Age, yr`        = msd(wave_age, 1),
    `Pubertal stage` = msd(pubertal_stage, 2),
    `Wear fraction`  = msd(wear_time_fraction, 2),
    .groups = "drop"
  ) |>
  kable(caption = "Table 1. Participant characteristics by behavioral phenotype (cluster) and visit. Cells are mean (SD) except counts. HOW TO READ: each row is one cluster at one visit; compare rows *within* a visit. Age and pubertal stage are near-identical across clusters within a visit — confirming clusters capture behavior, not maturation. 'Wear fraction' (proportion of the 21-day window with valid wear) is shown because it differs across clusters (Table 2) and is therefore adjusted for in all models.")
```

A confound check (Table 2) asks whether clusters merely reflect how much the device
was worn. Every Kruskal–Wallis test is significant, so wear differs across clusters;
because the absolute differences are modest (Supplementary Table S1: mean wear
fraction ≈ 0.83–0.94), wear is treated as a nuisance covariate rather than the signal.

```{r tbl-compliance}
qc_compliance |>
  transmute(Visit = wave_label, Metric = metric,
            `KW statistic` = round(statistic, 1), df = df,
            p = sapply(p_value, fmt_pf),
            `Confound?` = ifelse(confound_flag, "yes", "no")) |>
  kable(caption = "Table 2. Wear-compliance confound check (Kruskal–Wallis across clusters, per visit). HOW TO READ: a significant test means the metric (wear fraction, or number of valid days) is distributed differently across clusters — i.e. compliance is associated with phenotype. All tests are significant, which is why wear fraction enters every outcome model as a covariate; it does not by itself invalidate the phenotypes.")
```

```{r tbl-wear}
master |>
  group_by(Visit = wave_label, Cluster = cluster_aligned) |>
  summarise(N = dplyr::n(),
            `Valid days`           = msd(n_valid_days, 1),
            `Wear fraction`        = msd(wear_time_fraction, 2),
            `Valid-day dispersion` = msd(valid_day_dispersion, 2),
            .groups = "drop") |>
  kable(caption = "Table S1. Wear quality by cluster and visit; cells are mean (SD). HOW TO READ: this is the detail behind Table 2 — it shows the magnitude of compliance differences. 'Valid days' (out of 21) and 'wear fraction' are near-ceiling and similar across clusters; 'valid-day dispersion' (spread of valid days within the window) flags whether valid days bunch early/late. Similar values across clusters indicate compliance is a minor, adjustable nuisance.")
```

### Inference

Clusters are matched on age and pubertal stage within each visit (Table 1), so the
phenotypes reflect *behavior* rather than maturation — a precondition for interpreting
them as behavioral types. They do differ in wear compliance (Table 2), but only modestly
(Table S1). This is precisely why `wear_time_fraction` is carried as a covariate in every
outcome model that follows (Tables 7–10): it ensures the later behavioral associations
are not an artefact of how much each group wore the device.

## Behavioral phenotypes and their cross-wave (in)stability

`r n_clust` phenotypes were recovered at each visit. Tables 3a–b give their z-scored
profiles (0 = visit average); Figure 1 shows their separation in the embedding and
Figure 2 their mean diurnal shapes.

```{r tbl-zprofiles}
zlab <- c(
  z_steps_total_mean_daily         = "Steps (volume)",
  z_mets_mean_daily                = "METs (intensity)",
  z_steps_total_mvpa_fraction      = "MVPA fraction",
  z_steps_total_sedentary_fraction = "Sedentary fraction",
  z_min_slp_duration_mean          = "Sleep duration",
  z_min_slp_onset_mean             = "Sleep onset (+ later)",
  z_min_slp_onset_sd               = "Onset variability"
)
zprof <- function(w) {
  d <- master |> filter(wave_label == w) |>
    group_by(Cluster = cluster_aligned) |>
    summarise(N = dplyr::n(),
              across(any_of(names(zlab)), \(x) msd(x, 2)), .groups = "drop")
  pr <- intersect(names(zlab), names(d)); names(d)[match(pr, names(d))] <- zlab[pr]
  d
}
kable(zprof("yr2"), caption = "Table 3a. yr2 (~age 12) phenotype profiles: population z-scores, mean (SD). HOW TO READ: values are SDs from the visit mean (0 = average adolescent that visit; +1 = one SD above). Scan a column to see which cluster is high/low on that behavior. At yr2 the clusters separate mainly on SLEEP TIMING — note the large contrast in 'Sleep onset' and 'Onset variability' between clusters 0 and 1.")
kable(zprof("yr6"), caption = "Table 3b. yr6 (~age 16) phenotype profiles: population z-scores, mean (SD). HOW TO READ: as Table 3a. At yr6 the dominant contrast has shifted to PHYSICAL ACTIVITY — clusters now separate on Steps/METs/MVPA rather than sleep timing, illustrating that the defining behavioral axis changes with age.")
```

Note: `r n_noise_pv` participant-visit(s) with missing features were excluded from
clustering and do not appear in the UMAP or cluster-specific tables below; they are
retained in the feature-level models (Tables 7–8).

```{r fig-umap}
#| fig-cap: "Figure 1. Two-dimensional UMAP (Uniform Manifold Approximation and Projection) embedding of the behavioral feature space, coloured by aligned cluster, faceted by visit. Each point is one participant-visit; proximity reflects overall behavioral similarity across all features. Clear colour territories confirm the clusters occupy distinct regions of behavioral space. This is a 2-D projection for display only — clustering used the 10-D UMAP embedding. Participants excluded from clustering due to missing features are not shown (see Table P1)."
fig_umap
```

```{r fig-diurnal}
#| fig-cap: "Figure 2. Mean diurnal profiles by cluster (lines) with 95% confidence ribbons, per signal (rows) and visit (columns). Each curve is the cluster-mean of per-participant slot averages across valid days; ribbons are ±1.96 × SE across participants. HOW TO READ: the x-axis is hour of day (2-hour slots); vertical separation = different amounts of that behavior; horizontal shifts = different timing (e.g. a later-peaking activity curve, or a sleep curve shifted to later hours)."
#| fig-height: 7
if (!is.null(fig_diurnal)) fig_diurnal
```


```{r fig-onset-circular}
#| fig-cap: "Figure 2b. Circular (clock-face) distribution of mean sleep-onset time by cluster and visit. Midnight is at the top; the clock runs clockwise. Each bar represents the fraction of participants in that cluster whose average nightly sleep-onset hour falls in that 1-hour bin. Bars grow outward from the inner ring; a longer bar = more participants at that onset hour. Colours distinguish visits (blue = yr2, orange = yr6). Panels with no participants (e.g. a cluster absent at one visit after HDBSCAN alignment) are shown as empty rings. HOW TO READ: bars near 10 pm–midnight = early consistent onset; bars near 1–4 am = late or variable onset."
#| fig-height: 9
# # Pre-compute bin proportions so all panels use the same relative scale
# # (raw counts would make small clusters invisible relative to large ones)
# onset_counts <- master |>
#   dplyr::filter(!is.na(min_slp_onset_mean), !is.na(cluster_aligned)) |>
#   dplyr::mutate(
#     bin     = floor(min_slp_onset_mean %% 24),
#     Cluster = paste("Cluster", cluster_aligned),
#     Visit   = wave_label
#   ) |>
#   dplyr::count(Cluster, Visit, bin) |>
#   tidyr::complete(Cluster, Visit, bin = 0:23, fill = list(n = 0L)) |>
#   dplyr::group_by(Cluster, Visit) |>
#   dplyr::mutate(
#     total = sum(n),
#     prop  = dplyr::if_else(total > 0L, n / total, 0)
#   ) |>
#   dplyr::ungroup()
# 
# max_prop <- max(onset_counts$prop, na.rm = TRUE)
# inner    <- -max_prop * 0.65   # donut hole: pushes bars visually outward
# outer    <-  max_prop * 1.15
# 
# ggplot(onset_counts, aes(x = bin + 0.5, y = prop, fill = Visit)) +
#   geom_col(width = 0.9, colour = "white", linewidth = 0.05, alpha = 0.88) +
#   # start = 0 places x = 0 (midnight) at the top; direction = 1 = clockwise
#   coord_polar(start = 0, direction = 1) +
#   scale_x_continuous(
#     limits = c(0, 24),
#     breaks = c(0, 2, 4, 22),
#     labels = c("midnight", "2 am", "4 am", "10 pm")
#   ) +
#   scale_y_continuous(limits = c(inner, outer), expand = c(0, 0)) +
#   scale_fill_manual(values = c(yr2 = "#3B9AB2", yr6 = "#E07B39"), guide = "none") +
#   facet_grid(Cluster ~ Visit) +
#   labs(x = NULL, y = NULL) +
#   theme_minimal(11) +
#   theme(
#     axis.text.y        = element_blank(),
#     axis.ticks         = element_blank(),
#     axis.line          = element_blank(),
#     panel.grid.major.x = element_line(colour = "grey72", linewidth = 0.4),
#     panel.grid.minor.x = element_blank(),
#     panel.grid.major.y = element_blank(),   # suppress radial spoke lines
#     panel.grid.minor.y = element_blank(),
#     strip.text         = element_text(size = 9, face = "bold"),
#     panel.spacing      = unit(0.5, "lines"),
#     plot.margin        = margin(5, 20, 5, 20)
#   )
```

Critically, the phenotypes are **age-specific states, not stable traits.** Within-wave
clustering was highly reproducible (bootstrap Adjusted Rand Index > 0.9, clustering
log), but the cross-wave alignment is heterogeneous (Table 4):
`r n_preserved` of `r nrow(alignment)` phenotypes are well preserved from yr2 to yr6
(cosine similarity up to `r round(al_max, 2)`), whereas cluster `r al_low` is only
weakly preserved (`r round(al_min, 2)`). A low similarity
means the centroid that best matches across ages is nonetheless quite different — the
phenotype does not carry over.

```{r tbl-alignment}
alignment |>
  transmute(`Aligned cluster` = aligned_cluster,
            `From` = ref_wave, `To` = target_wave,
            `Centroid cosine similarity` = round(cosine_similarity, 3)) |>
  kable(caption = "Table 4. Cross-wave cluster alignment (Supp. Table S3). HOW TO READ: each aligned label was matched yr2→yr6 by maximizing centroid similarity; the score (1 = identical centroid, 0 = orthogonal) says how well that phenotype is preserved across age. High values (≳0.8) license treating the label as the same phenotype at both visits; the low value for one cluster is the quantitative basis for NOT pooling cluster effects across waves in the outcome models.")
```

### Inference

Two facts combine into one conclusion. The within-visit clusters are reproducible (high
bootstrap ARI), yet (i) the discriminating axis *changes with age* — sleep timing at yr2
(Table 3a) versus physical activity at yr6 (Table 3b) — and (ii) one phenotype is poorly
preserved across visits (cluster `r al_low`, similarity `r round(al_min, 2)`; Table 4).
The cluster labels are therefore **age-specific, not the same construct over time.** This
is the rationale for two deliberate downstream choices: the primary outcome analysis uses
continuous features whose meaning is constant across visits (Table 7), and cluster–outcome
effects are estimated *within* visit rather than pooled across age (Table 9). It also makes
a falsifiable prediction — phenotype membership should not persist — which the next section
tests directly.

## Phenotype transitions and regularity trajectories

Consistent with weak alignment, cluster membership does not carry across age. Among the
`r transitions$n_complete` participants clustered at both visits, the transition matrix
(Table 5) does not depart from independence (χ² `r fmt_p(tr_p)`): a participant's yr2
phenotype does not predict their yr6 phenotype. Figure 3 shows the corresponding flows.

```{r tbl-transitions}
transitions$counts |>
  tibble::rownames_to_column("yr2 cluster") |>
  kable(caption = "Table 5. Cluster transition counts (rows = yr2 cluster, columns = yr6 cluster). HOW TO READ: each cell counts participants moving from a yr2 phenotype (row) to a yr6 phenotype (column); the diagonal would be large if phenotypes were stable. Here every row distributes similarly across columns, i.e. the destination is roughly independent of the origin — phenotypes reshuffle between ages 12 and 16.")
```

```{r fig-alluvial}
#| fig-cap: "Figure 3. Alluvial flows of cluster membership from yr2 to yr6. HOW TO READ: band width is proportional to the number of participants; crossing/splitting bands show movement between phenotypes. The heavy crossing visualizes the non-significant transition test — most participants change phenotype across age."
fig_alluvial
```

Modelling the *path* of behavioral regularity directly (Group-Based Trajectory
Modeling [GBTM] of sleep-onset variability) selected **`r gbtm$best_ng` trajectory
groups** by the Bayesian Information Criterion (BIC; Table 6); their characterization (Table 6b) yields three developmentally
distinct sub-populations: a large *consistently regular* group, a minority whose
regularity *worsens* with age, and a small *persistently irregular* group. Mean
posterior assignment probability was `r round(gbtm$quality$mean_max_posterior, 2)`
(`r gbtm$quality$pct_confident_0.8`% of participants assigned at > 0.8 probability),
so memberships are used as soft probabilities downstream.

```{r tbl-gbtm}
gbtm$bic |>
  mutate(BIC = round(BIC, 1)) |>
  kable(caption = "Table 6. Trajectory-model fit. HOW TO READ: BIC trades fit against complexity (lower = better); the minimum identifies the supported number of latent regularity paths.",
        col.names = c("N groups", "BIC"))
```

```{r tbl-gbtm-classes}
mj <- dplyr::left_join(master, dplyr::select(gbtm$membership, participant_id, traj_group),
                       by = "participant_id")
tab6b <- mj |>
  filter(!is.na(traj_group)) |>
  group_by(Group = traj_group) |>
  summarise(N = dplyr::n_distinct(participant_id),
            `Onset SD, yr2 (h)` = msd(min_slp_onset_sd[wave_label == "yr2"], 2),
            `Onset SD, yr6 (h)` = msd(min_slp_onset_sd[wave_label == "yr6"], 2),
            .groups = "drop")
tab6b <- dplyr::left_join(
  tab6b,
  dplyr::transmute(gbtm$class_summary, Group = role, `%` = pct,
                   `Change/yr (h)` = round(slope_per_year, 3)),
  by = "Group")
kable(tab6b, caption = "Table 6b. Trajectory-group characterization — sleep-onset variability (SD of onset clock time, in hours; lower = more regular bedtimes). Cells are mean (SD). HOW TO READ: compare the yr2 and yr6 columns within a row to see each group's developmental change, and the 'Change/yr' column for its slope. The 'worsening' group's onset SD rises markedly from yr2 to yr6; the 'regular' group is low and flat; the 'irregular' group is high at both ages.")
```

### Inference

The non-significant transition test (Table 5, Figure 3) confirms the prediction from the
previous section: a participant's phenotype at age 12 carries essentially no information
about their phenotype at 16 — these are states, not traits. Because cluster *trajectories*
are thus uninformative, we instead model the path of a single consistently-defined
behavior, sleep-onset regularity. The trajectory model isolates a minority "worsening"
group (Table 6b) whose deteriorating regularity is a genuinely longitudinal construct
(unlike the per-visit clusters). That group is carried forward as a predictor of mental
health (Table 8), connecting this descriptive result to the primary outcome analysis.

## Primary analysis: behavioral features, age, and mental health

The pre-specified primary models regress diagnosis burden on each z-scored feature and
its interaction with age (Table 7). `r n_sig_feat` of four features show a
significant, FDR-surviving **feature × age** interaction: sleep-onset variability
(β = `r signif(on_b, 2)`, FDR `r fmt_pf(on_fdr)`), daily steps
(β = `r signif(st_b, 2)`, FDR `r fmt_pf(st_fdr)`), and MVPA fraction
(β = `r signif(mv_b, 2)`, FDR `r fmt_pf(mv_fdr)`).
`r if (!is.na(on_b) && on_b > 0 && !is.na(st_b) && st_b < 0) "The positive sign for sleep irregularity and negative sign for the activity features mean that," else "The direction of these effects means that,"` **as adolescents move from 12 to 16, greater sleep irregularity becomes increasingly associated with higher psychiatric burden, while higher physical activity becomes increasingly protective.** Sleep duration shows no such moderation.

**Model specification** (fitted separately for each of the four pre-specified features;
BH-FDR correction across the four `z_feature:wave_age` p-values):

$$\texttt{mh\_burden\_present} \sim z\_\text{feature} \times \texttt{wave\_age} + \texttt{sex} + \texttt{pubertal\_stage} + \texttt{wear\_time\_fraction} + (1 \mid \texttt{participant\_id})$$

```{r tbl-features}
flab <- c(z_min_slp_onset_sd = "Sleep-onset variability",
          z_steps_total_mean_daily = "Daily steps",
          z_steps_total_mvpa_fraction = "MVPA fraction",
          z_min_slp_duration_mean = "Sleep duration")
mh_feature_summary |>
  transmute(Feature = flab[feature],
            `β main (SE)` = sprintf("%.3f (%.3f)", beta_main, se_main),
            `p main`      = sapply(p_main, fmt_pf),
            `β ×age (SE)` = sprintf("%.3f (%.3f)", beta_x_age, se_x_age),
            `p ×age`      = sapply(p_x_age, fmt_pf),
            `FDR p ×age`  = sapply(p_x_age_fdr, fmt_pf)) |>
  kable(caption = "Table 7. PRIMARY models — feature × age associations with diagnosis burden (mixed models adjusted for sex, pubertal stage, wear fraction; participant random intercept). Features are within-wave z-scores, so β is the change in burden per +1 SD of the behavior. HOW TO READ: the 'β ×age' column is the key test — it is how much the feature–burden association changes per year of age. A positive ×age (sleep-onset variability) means the harmful association strengthens with age; a negative ×age (steps, MVPA) means the protective association strengthens with age. 'FDR p' is the Benjamini–Hochberg-adjusted significance across the four features. (Main-effect βs are conditional on age = 0 and are not interpreted directly.)")
```

### Inference

This is the study's primary result: higher physical activity becomes increasingly
*protective*, and greater sleep irregularity increasingly *harmful*, for psychiatric
burden as adolescents move from 12 to 16 (Table 7). Using features rather than clusters is
deliberate — they are the only behavioral predictors whose meaning is constant across
visits, so the feature × age interaction is interpretable where a cluster × age term would
not be (cf. cross-wave instability, Table 4). The two sections that follow corroborate this
result from independent angles: the worsening-regularity trajectory group (Table 8) and the
within-visit low-activity cluster (Table 9). The sleep-duration null is informative — the
effect is specific to the *timing/regularity* of sleep, not its quantity.

## Convergent evidence: regularity-trajectory group

Replacing the within-wave features with the longitudinally-coherent trajectory groups
gives a convergent result (Table 8): the *worsening-regularity* group shows a steeper
age-related rise in burden than the regular reference (probability × age
β = `r signif(trw_b, 2)`, `r fmt_p(trw_p)`), echoing the sleep-irregularity finding
above from an independent modelling route.

**Model specification** (posterior probabilities as continuous predictors;
reference group = "regular"):

$$\texttt{mh\_burden\_present} \sim \texttt{prob\_worsening} \times \texttt{wave\_age} + \texttt{prob\_irregular} \times \texttt{wave\_age} + \texttt{sex} + \texttt{pubertal\_stage} + \texttt{wear\_time\_fraction} + (1 \mid \texttt{participant\_id})$$

```{r tbl-traj}
mh_traj_model$table |>
  transmute(Term = term, `Estimate (SE)` = sprintf("%.3f (%.3f)", Estimate, `Std. Error`),
            t = round(`t value`, 2), p = sapply(`Pr(>|t|)`, fmt_pf)) |>
  kable(caption = sprintf("Table 8. Trajectory-group model of diagnosis burden (n = %s observations). Predictors are posterior probabilities of the 'worsening' and 'irregular' groups (reference = regular), each interacted with age. HOW TO READ: the ':wave_age' rows test whether a group's burden gap vs. the regular group widens with age; a positive, significant 'prob_worsening:wave_age' means the worsening-regularity group accrues psychiatric burden faster across adolescence.", format(mh_traj_model$n_obs, big.mark = ",")))
```

### Inference

The worsening-regularity group accrues psychiatric burden faster with age (Table 8) — an
independent replication of the sleep-onset-variability × age effect from the primary
feature models (Table 7), now via a person-level latent-trajectory construct rather than a
per-visit measurement. Because a continuous-feature analysis and a latent-class analysis
converge on the same conclusion, an analytic-artefact explanation is unlikely; the
trajectory framing additionally supplies a temporal ordering — a *worsening path* of
regularity accompanies rising burden across adolescence.

## Within-wave cluster associations

Estimating cluster effects nested within visit (no cross-wave pooling) localizes the
signal (Table 9): the most activity-divergent phenotype at yr6 (cluster `r yr6_cl_id`)
has elevated burden **at yr6** (β = `r signif(cl6_b, 2)`, `r fmt_p(cl6_p)`) but
**not at yr2** (`r fmt_p(cl2_p)`). This is consistent with the primary finding —
activity matters for mental health by mid-adolescence — and shows why a pooled
cluster × age term (which the weak alignment of cluster `r al_low` would have rendered
uninterpretable) is unnecessary here.

**Model specification — primary burden outcome (LMM):**

$$\texttt{mh\_burden\_present} \sim \texttt{wave\_label} + \texttt{wave\_label}{:}\texttt{cluster\_aligned} + \texttt{sex} + \texttt{pubertal\_stage} + \texttt{wear\_time\_fraction} + (1 \mid \texttt{participant\_id})$$

Clusters nested within visit; reference cluster = 0. Participants without a cluster
assignment (`r n_noise_pv` participant-visit[s]) are excluded from this model.

```{r tbl-cluster}
mh_models$burden |>
  transmute(Term = term, `Estimate (SE)` = sprintf("%.3f (%.3f)", Estimate, `Std. Error`),
            df = round(df), t = round(`t value`, 2), p = sapply(`Pr(>|t|)`, fmt_pf)) |>
  kable(caption = "Table 9. Within-wave (visit-nested) cluster model of diagnosis burden. 'wave:cluster' terms are each cluster's contrast vs. the reference cluster (0) *within* that visit. HOW TO READ: read the four 'wave_label..:cluster_aligned..' rows as wave-specific phenotype effects — e.g. 'wave_labelyr6:cluster_aligned1' is the burden difference for cluster 1 vs. cluster 0 at yr6. No term assumes a cluster means the same thing across visits.")
```

**Model specification — secondary any-depression outcome (Generalized Linear Mixed
Model [GLMM], logistic link):**

$$\texttt{mh\_any\_depression} \sim \texttt{wave\_label} + \texttt{wave\_label}{:}\texttt{cluster\_aligned} + \texttt{sex} + \texttt{pubertal\_stage} + \texttt{wear\_time\_fraction} + (1 \mid \texttt{participant\_id}), \quad \text{family = binomial}$$

```{r tbl-depression, results='asis'}
if (!is.null(mh_models$depression)) {
  mh_models$depression |>
    transmute(Term = term,
              `OR (SE)` = sprintf("%.3f (%.3f)", exp(Estimate), `Std. Error`),
              z = round(`z value`, 2),
              p = sapply(`Pr(>|z|)`, fmt_pf)) |>
    kable(caption = "Table 9b. Secondary outcome: any-depression (logistic GLMM, odds ratios). Same wave-nested specification as Table 9. OR > 1 indicates higher odds of any current depressive diagnosis.")
}
```

### Inference

The visit-nested model localizes the activity–mental-health link in time: the low-activity
phenotype carries elevated burden at age 16 but not age 12 (Table 9), mirroring the
negative steps/MVPA × age interactions of the primary models (Table 7). This is the third
convergent line of evidence (with Tables 7 and 8). It also demonstrates the payoff of the
wave-nested specification — the effect is cleanly attributable to the age-16 phenotype,
whereas a pooled cluster × age term would have been confounded by this cluster's poor
cross-wave alignment (Table 4).

## Metabolic outcomes

Across all three labs, no cluster term was significant; metabolic variation tracked age,
sex, and pubertal stage instead (Table 10). The behavioral–metabolic link is weak here,
though the metabolic sample is small (labs are sparsely available).

**Model specification** (same wave-nested structure as Table 9, fitted separately for
each metabolic lab):

$$\texttt{lab\_value} \sim \texttt{wave\_label} + \texttt{wave\_label}{:}\texttt{cluster\_aligned} + \texttt{sex} + \texttt{pubertal\_stage} + \texttt{wear\_time\_fraction} + (1 \mid \texttt{participant\_id})$$

```{r tbl-metabolic, results='asis'}
metlab <- c(ph_y_bld__rslt__hdl_qnt = "HDL cholesterol",
            ph_y_bld__rslt__chol_qnt = "Total cholesterol",
            ph_y_bld__rslt__a1c_qnt = "Hemoglobin A1c")
if (length(metabolic_models) > 0) {
  for (nm in names(metabolic_models)) {
    cat("\n\n**", ifelse(nm %in% names(metlab), metlab[[nm]], nm), "**\n\n", sep = "")
    print(metabolic_models[[nm]] |>
            transmute(Term = term, `Estimate (SE)` = sprintf("%.3f (%.3f)", Estimate, `Std. Error`),
                      t = round(`t value`, 2), p = sapply(`Pr(>|t|)`, fmt_pf)) |>
            kable())
  }
  cat("\n\n*Table 10. Visit-nested cluster models for each metabolic lab (adjusted for sex, pubertal stage, wear fraction; participant random intercept). HOW TO READ: as Table 9. Look for significant 'wave:cluster' rows (behavioral-phenotype effects) versus the covariate rows — here the significant terms are age (wave) and sex, not cluster, indicating maturation/sex rather than behavior drives these labs.*\n\n")
}
```

### Inference

In contrast to mental health, behavioral phenotype explains no metabolic variation
(Table 10); the labs are governed by maturation (age, pubertal stage) and sex. This
dissociation is itself informative — the wearable behavioral signal is *outcome-specific*,
clearly present for psychiatric burden (Tables 7–9) but not for these cardiometabolic
markers. Whether that reflects a genuinely weaker behavioral–metabolic coupling at these
ages, or simply the small and sparsely-measured metabolic sample, cannot be resolved here
(see Limitations).

## Multivariate behavioral–health axis

**Model specification:**
- *Canonical Correlation Analysis (CCA):* `stats::cancor(Z_features, outcomes)` where
  `Z_features` is the matrix of all z-scored behavioral features and `outcomes` is the
  combined mental-health and metabolic outcome set.
- *Cluster MANOVA:* `manova(Z_features ~ cluster_aligned * wave_label)` testing whether
  clusters separate the full feature space.

A CCA between the full feature set and the combined outcomes (burden + labs) found a
moderate first canonical correlation (`r round(cc1, 2)`; Figure 4), and the cluster
Multivariate Analysis of Variance (MANOVA) was marginal (Pillai `r fmt_p(man_p)`) —
suggestive of a shared behavioral–health axis but not conclusive at the current
(lab-limited) sample size.

```{r fig-canonical}
#| fig-cap: "Figure 4. First canonical variate pair. HOW TO READ: each point is a participant scored on the single best-correlated linear combination of behavioral features (x) and of outcomes (y); the fitted line's tightness reflects the canonical correlation. A clear positive slope indicates a coherent axis along which 'more of this behavior pattern' co-varies with 'more of this outcome pattern'."
if (!is.null(fig_canonical)) fig_canonical
```

### Inference

Taking all features and outcomes jointly, there is a moderate shared axis (canonical
r `r round(cc1, 2)`; Figure 4) but only a marginal omnibus cluster effect (MANOVA Pillai
`r fmt_p(man_p)`). Read alongside the outcome-specific models, this shared axis is carried
mainly by the mental-health side (Tables 7–9), since the metabolic labs contributed little
(Table 10). The multivariate result is therefore consistent with — but weaker than — the
targeted analyses, and is constrained by the lab-limited sample.

# Discussion

This study derived data-driven behavioral phenotypes from 2-hour-resolution Fitbit
activity and sleep in adolescents and related them to mental-health and metabolic outcomes
across two visits. Three conclusions follow, each supported by convergent evidence above.

**Behavioral phenotypes are age-specific states, not stable traits.** Within-visit
clustering was highly reproducible, yet the organizing axis of behavior shifted from sleep
timing at age 12 to physical-activity level at age 16 (Tables 3a–b),
`r nrow(alignment) - n_preserved` of `r nrow(alignment)` phenotypes were poorly
preserved across age (lowest alignment similarity `r round(al_min, 2)`; Table 4), and
membership did not carry over (non-significant transitions; Table 5, Figure 3). A single-
visit "phenotype" should therefore not be read as an enduring trait, and cross-sectional
snapshots can mischaracterize how these behaviors relate to health across development.

**Physical activity becomes protective, and sleep irregularity harmful, for mental health
by mid-adolescence.** The pre-specified feature × age models — the analyses least
vulnerable to label instability — showed FDR-significant developmental moderation, with
higher activity increasingly associated with lower psychiatric burden and greater sleep-
onset variability with higher burden as adolescents aged from 12 to 16 (Table 7). The same
pattern arose independently from the worsening-regularity trajectory group (Table 8) and
the visit-nested low-activity cluster (Table 9). This triangulation across continuous-
feature, latent-trajectory, and within-visit-cluster analyses makes an artefactual
explanation unlikely and indicates a developmentally timed relationship that intensifies
through adolescence.

**Metabolic outcomes were driven by maturation, not behavior.** Routine cardiometabolic
labs tracked age, sex, and pubertal stage with no detectable behavioral-phenotype
contribution (Table 10) and only a moderate, non-conclusive multivariate signal
(Figure 4). The wearable behavioral signal here is thus outcome-specific — robust for
mental health, absent for these metabolic markers.

Methodologically, the analysis illustrates a general pitfall and remedy for longitudinal
clustering. Independently-derived per-visit clusters were re-labelled by centroid matching,
but **best-match alignment does not guarantee that a label denotes the same construct over
time**; pooling such labels into a cluster × age interaction can manufacture or distort
developmental effects. Quantifying alignment similarity (Table 4) made the problem visible,
and basing primary inference on consistently-defined continuous features — while reserving
clusters for within-visit description — yielded interpretable, convergent results. We
recommend this sequence (quantify cross-wave alignment, then prefer consistently-defined
predictors for longitudinal inference) for wearable-phenotyping studies generally.

## Limitations

- **Two visits only.** With measurements at ages ~12 and ~16, every longitudinal construct
  is effectively a two-point contrast: trajectory shapes are linear by construction and
  transitions are a single step. A third visit (yr4) would permit non-linear trajectories
  and more powerful transition modelling.
- **Signal scope.** Resting heart rate was excluded as unreliable, and sleep was derived
  from minutes-of-sleep per epoch rather than staged sleep, so onset and efficiency are
  approximations pending raw per-minute data.
- **Circadian resolution.** Circadian indices (IS, IV, cosinor amplitude/acrophase, L5/M10)
  are estimated from 12 two-hour bins per day. Single-harmonic cosinor and IS are
  statistically coarse at this resolution; results are interpreted as approximate circadian
  phenotypes rather than precise rhythm parameters.
- **Metabolic power.** Blood labs were sparsely available; the metabolic null is
  power-limited rather than strong evidence of no effect, and it caps the multivariate
  sample.
- **Observational design.** Despite adjustment for wear compliance, sex, age, and pubertal
  stage, residual confounding cannot be excluded and associations are not causal.
- **Modelling approximations.** Diagnosis burden (a count) was modelled with a linear mixed
  model; feature main effects are conditional on uncentered age and are not interpreted;
  only the pre-specified primary feature interactions were FDR-corrected, so cluster and
  metabolic contrasts are exploratory. Trajectory-group membership was carried as soft
  probabilities to mitigate, but not eliminate, classification uncertainty.
- **Generalizability.** Analyses used a curated subset of participants with complete sleep
  and step data at both visits; compliant wearers may differ from the broader cohort.

---

*Reproducible report generated from the `targets` pipeline (`_targets.R`); all tables
and figures are read from the pipeline store. Rebuild with `targets::tar_make()` and
re-render.*
