---
title: "Biological Rhythm Extraction from Adolescent Fitbit Data"
subtitle: "Replication of Yan & Doryab (2022) · ABCD Cohort · 3 Waves"
author: "Biplabendu Das"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    theme: lumen
    fig-width: 10
    fig-height: 5
    embed-resources: true
    code-fold: show
    code-tools: true
execute:
  warning: false
  message: false
  cache: true
---

## Overview

This document applies the computational framework from **Yan & Doryab (2022)** — *"Towards a Computational Framework for Automated Discovery and Modeling of Biological Rhythms from Wearable Data Streams"* (IntelliSys 2021, LNNS 296, pp. 643–661) — to the ABCD Study Fitbit cohort.

The framework identifies biological rhythms in wearable sensor data through three sequential stages: (1) periodicity detection using **four complementary methods** — FFT, Chi-squared periodogram, Lomb-Scargle periodogram, and autocorrelation (ACF); a fifth method, a cyclic HMM approximation (`depmixS4`), is implemented but temporarily disabled for speed; (2) change-point detection within each confirmed period using AutoNOM and Rbeast; and (3) feature extraction for longitudinal growth curve modeling.

**Adaptation notes for this dataset:**

- Data resolution: 2-hour slots (12 slots/day), not minute-level as in the paper
- Signals: `steps_total`, `min_slp`, `mets` (HR excluded — unreliable for consumer Fitbit)
- Three waves: `ses-02A` (yr2, ~age 12), `ses-04A` (yr4, ~age 14), `ses-06A` (yr6, ~age 16)
- Cohort: 2,919 participants (ID-filtered for ≥2 weekends of data at yr2 and yr6)
- **Period consensus: ≥75% of non-NA methods must detect a 24-hr period (±4 hr).** With all 4 active methods reporting, this means ≥3 of 4 must agree. The paper required all 3 of its methods; we use a proportional threshold to stay robust at the coarser 2-hr resolution
- **CyHMM currently disabled (temporary):** the original Pierson et al. (2018) Python CyHMM is Python-2 only and depends on a deprecated `pomegranate` API; the `depmixS4` approximation that replaced it is too slow, so it is switched off for now. Periodicity detection runs on the four frequency-/lag-domain methods (FFT, Chi-squared, Lomb-Scargle, ACF). The CyHMM code is retained (commented) for re-enabling
- AutoNOM and Rbeast applied to full wave series (~252 time points), not per-day, due to resolution constraints

---

## Setup

```{r setup}
test = TRUE

library(arrow)
library(tidyverse)
library(lubridate)
library(imputeTS)
library(lomb)         # Lomb-Scargle periodogram (handles missing/irregular data)
library(depmixS4)     # cyclic HMM approximation (HMM with constrained transitions)
library(Rbeast)       # change-point detection (fallback + complement)
library(lme4)
library(lmerTest)
library(knitr)        # kable tables (no gt/V8 dependency)

# lme4/lmerTest load MASS, whose select() masks dplyr::select() on the search path.
# Reclaim the dplyr verbs we rely on so the pipeline behaves regardless of load order.
select <- dplyr::select
filter <- dplyr::filter

# Lightweight table helper — matches the existing index.qmd convention and avoids
# the gt → juicyjuice → V8 toolchain (V8 won't compile under the Homebrew-GCC Makevars).
# A trailing `subtitle` is appended to the caption in parentheses.
ktab <- function(df, caption = NULL, subtitle = NULL, digits = 2) {
  cap <- caption
  if (!is.null(subtitle)) cap <- paste0(caption, " — ", subtitle)
  knitr::kable(df, caption = cap, digits = digits, format.args = list(big.mark = ","))
}

# ── Resolution constants (2-hr slots) ────────────────────────────────────────
SLOT_HOURS        <- 2L           # hours per slot
SLOTS_PER_DAY     <- 12L          # 24 hr / 2 hr
SLOT_STARTS       <- seq(0, 22, by = 2)  # slot start hours: 0, 2, …, 22
SLOT_MIDPOINTS    <- SLOT_STARTS + 1     # midpoint hours for cosinor: 1, 3, …, 23

# ── Validity thresholds ───────────────────────────────────────────────────────
MIN_VALID_SLOTS   <- 8L           # ≥8 of 12 slots present → valid day
MIN_VALID_DAYS    <- 14L          # ≥14 valid days per wave → keep participant-wave
MAX_IMPUTE_GAP    <- 3L           # max consecutive missing slots to impute
SMA_K             <- 2L           # SMA window: ±2 slots = ±4 hr

# ── Periodicity detection ─────────────────────────────────────────────────────
# Active methods: FFT, chi-squared periodogram, Lomb-Scargle, ACF.
# (depmixS4 cyclic HMM is temporarily disabled — too slow; see §2.5.)
PERIOD_MIN_HR     <- 12           # minimum period to test (hours)
PERIOD_MAX_HR     <- 48           # maximum period to test (hours)
PERIOD_STEP_HR    <- 0.5          # step size for chi-sq scan (hours)
CHI_ALPHA         <- 0.01         # significance threshold
LS_ALPHA          <- 0.01         # Lomb-Scargle significance threshold
PERIOD_TOL_HR     <- 4            # ±4 hr tolerance for method agreement
TARGET_PERIOD_HR  <- 24           # expected circadian period
N_METHODS         <- 4            # FFT, chi-sq, Lomb-Scargle, ACF (CyHMM disabled)
CONSENSUS_FRAC    <- 0.75         # ≥75% of non-NA methods must agree → ≥3 of 4 confirmed

# ── AutoNOM / change-point ────────────────────────────────────────────────────
KMAX_RANGE        <- 3:5          # kmax values for MAPE-based selection
MMAX              <- 3L           # max frequency components per AutoNOM segment

# ── Non-parametric circadian features ────────────────────────────────────────
M10_SLOTS         <- 5L           # 10-hr most active window (5 × 2 hr)
L5_SLOTS          <- 3L           # 6-hr least active window (approx L5; 3 × 2 hr)

# ── Paths ─────────────────────────────────────────────────────────────────────
RAW_PARQUET  <- "dev/data/fitbit-summaries/activity_120m.parquet"
ID_FILE      <- "data/ids_fitbit_v02_v06-sleep_steps_data.csv"
SIGNALS      <- c("steps_total", "min_slp", "mets")
SESSIONS     <- c("ses-02A", "ses-04A", "ses-06A")
WAVE_LABELS  <- c("ses-02A" = "yr2", "ses-04A" = "yr4", "ses-06A" = "yr6")
WAVE_AGES    <- c("yr2" = 12, "yr4" = 14, "yr6" = 16)
```

---

## Stage 0: Data Summary

Examine the raw parquet file before any filtering. This section is the entry point for understanding data structure, session coverage, and signal missingness.

```{r data-load}
raw <- read_parquet(RAW_PARQUET)
```

### Column schema

```{r schema}
tibble(
  column = names(raw),
  type   = map_chr(raw, \(x) paste(class(x), collapse = " / "))
) |>
  rename(Column = column, `R type` = type) |>
  ktab("Raw parquet column schema")
```

### Participant and session coverage

```{r coverage}
raw |>
  group_by(session_id) |>
  summarise(
    n_participants = n_distinct(participant_id),
    n_participant_days = n_distinct(paste(participant_id, day)),
    median_days = median(
      (count(pick(participant_id, day), participant_id) |> pull(n)),
      na.rm = TRUE
    ),
    .groups = "drop"
  ) |>
  arrange(session_id) |>
  ktab("Participants and days per session (unfiltered)")
```

```{r days-per-participant}
days_per_pw <- raw |>
  filter(session_id %in% SESSIONS) |>
  group_by(participant_id, session_id) |>
  summarise(n_days = n_distinct(day), .groups = "drop")

days_per_pw |>
  group_by(session_id) |>
  summarise(
    n_participants = n(),
    min    = min(n_days),
    q25    = quantile(n_days, 0.25),
    median = median(n_days),
    q75    = quantile(n_days, 0.75),
    max    = max(n_days),
    .groups = "drop"
  ) |>
  ktab("Distribution of days per participant-wave (3 sessions)", digits = 1)
```

### Signal missingness

Non-wear is identified as slots where `min_total == 0` (no recorded minutes). Signal-level NA indicates true missing data on top of non-wear.

```{r missingness}
raw |>
  filter(session_id %in% SESSIONS) |>
  mutate(non_wear = (is.na(min_total) | min_total == 0)) |>
  summarise(
    across(
      all_of(c("min_total", SIGNALS)),
      list(
        pct_na      = \(x) round(mean(is.na(x)) * 100, 1),
        pct_nonwear = \(x) round(mean(is.na(x) | (exists("non_wear") & non_wear)) * 100, 1)
      )
    )
  ) |>
  pivot_longer(everything(), names_to = c("signal", ".value"), names_sep = "_pct_") |>
  rename(Signal = signal, `% NA` = na, `% NA or non-wear` = nonwear) |>
  ktab("Signal missingness across all 3 sessions (%)")
```

```{r missingness-by-session}
raw |>
  filter(session_id %in% SESSIONS) |>
  mutate(non_wear = (is.na(min_total) | min_total == 0)) |>
  group_by(session_id) |>
  summarise(
    across(
      all_of(SIGNALS),
      \(x) round(mean(is.na(x)) * 100, 1),
      .names = "{.col}_pct_na"
    ),
    pct_non_wear = round(mean(non_wear) * 100, 1),
    .groups = "drop"
  ) |>
  ktab("Signal missingness by session (%)")
```

### Valid-day preview

A day is valid if ≥8 of 12 slots have `min_total > 0` (≤ 8 hours non-wear). This mirrors the existing Python pipeline threshold.

```{r valid-day-preview}
raw |>
  filter(session_id %in% SESSIONS) |>
  group_by(participant_id, session_id, day) |>
  summarise(
    n_slots        = n(),
    n_valid_slots  = sum(!is.na(min_total) & min_total > 0),
    is_valid       = n_valid_slots >= MIN_VALID_SLOTS,
    .groups = "drop"
  ) |>
  group_by(session_id) |>
  summarise(
    total_participant_days = n(),
    pct_valid_days = round(mean(is_valid) * 100, 1),
    median_valid_slots = median(n_valid_slots),
    .groups = "drop"
  ) |>
  ktab("Day validity preview by session")
```

---

## Stage 1: Preprocessing

Following the paper's Stage 1 (Section 3), we: (1) apply the participant ID filter, (2) classify non-wear slots and invalid days, (3) apply SMA imputation to isolated missing values within valid days, and (4) retain only participant-waves with ≥14 valid days.

```{r preprocess}
valid_ids <- read_csv(ID_FILE, show_col_types = FALSE) |> pull(participant_id)

if (test) {
  valid_ids = sample(valid_ids, 500, replace = FALSE)
}

# Apply ID and session filters; label waves
dat <- raw |>
  filter(
    session_id     %in% SESSIONS,
    participant_id %in% valid_ids
  ) |>
  mutate(
    wave      = recode(session_id, !!!WAVE_LABELS),
    slot_hour  = hour(start),         # slot start hour (0, 2, 4, …, 22)
    is_weekend = wday(start, week_start = 1) >= 6  # Sat=6, Sun=7 (lubridate)
  )

n_before <- n_distinct(paste(dat$participant_id, dat$wave))
```

```{r non-wear}
dat <- dat |>
  mutate(non_wear = is.na(min_total) | min_total == 0)
```

```{r day-validity}
day_validity <- dat |>
  group_by(participant_id, wave, day) |>
  summarise(
    n_slots       = n(),
    n_valid_slots = sum(!non_wear),
    is_valid      = n_valid_slots >= MIN_VALID_SLOTS,
    .groups = "drop"
  )

wave_validity <- day_validity |>
  group_by(participant_id, wave) |>
  summarise(
    n_valid_days = sum(is_valid),
    passes       = n_valid_days >= MIN_VALID_DAYS,
    .groups = "drop"
  )
```

```{r preprocess-summary}
wave_validity |>
  group_by(wave) |>
  summarise(
    n_total          = n(),
    n_pass           = sum(passes),
    pct_pass         = round(mean(passes) * 100, 1),
    median_valid_days = median(n_valid_days),
    .groups = "drop"
  ) |>
  ktab(
    "Participant-wave counts after validity filtering",
    subtitle = glue::glue("Threshold: ≥{MIN_VALID_DAYS} valid days (≥{MIN_VALID_SLOTS} slots/day)")
  )
```

```{r filter-and-impute}
# Keep only valid days from participant-waves that pass the threshold
dat_valid <- dat |>
  semi_join(day_validity  |> filter(is_valid), by = c("participant_id", "wave", "day")) |>
  semi_join(wave_validity |> filter(passes),   by = c("participant_id", "wave"))

# SMA imputation: set non-wear slots to NA, then fill short gaps within each wave
# Imputation is applied only to valid-day slots; non-wear slots stay NA
dat_imputed <- dat_valid |>
  group_by(participant_id, wave) |>
  arrange(day, slot_hour, .by_group = TRUE) |>
  mutate(
    steps_total = if_else(non_wear, NA_real_, steps_total),
    min_slp     = if_else(non_wear, NA_real_, min_slp),
    mets        = if_else(non_wear, NA_real_, mets),
    steps_total = na_ma(steps_total, k = SMA_K, weighting = "simple"),
    min_slp     = na_ma(min_slp,     k = SMA_K, weighting = "simple"),
    mets        = na_ma(mets,        k = SMA_K, weighting = "simple")
  ) |>
  ungroup()
```

```{r imputation-summary}
bind_rows(
  dat_valid   |> summarise(across(all_of(SIGNALS), \(x) mean(is.na(x))), stage = "pre-imputation"),
  dat_imputed |> summarise(across(all_of(SIGNALS), \(x) mean(is.na(x))), stage = "post-imputation")
) |>
  pivot_longer(all_of(SIGNALS), names_to = "signal", values_to = "pct_na") |>
  mutate(pct_na = round(pct_na * 100, 2)) |>
  pivot_wider(names_from = stage, values_from = pct_na) |>
  ktab("Residual NA (%) before and after SMA imputation")
```

---

## Stage 2: Periodicity Detection

We apply **four complementary methods** to the concatenated wave time series for each participant × wave × signal combination, then confirm a 24-hr period if **≥75% of the non-NA methods** detect it within ±4 hours of the target (i.e. ≥3 of 4 when all methods report).

| # | Method | Function | Domain | Handles gaps | Status |
|---|--------|----------|--------|--------------|--------|
| 1 | Fast Fourier Transform | `run_fft()` | Frequency | Interpolated | active |
| 2 | Chi-squared periodogram (Q_P) | `chi_sq_periodogram()` | Variance-ratio | NA-dropped | active |
| 3 | Lomb-Scargle periodogram | `run_lomb()` | Frequency | Native | active |
| 4 | Autocorrelation (ACF) | `run_acf()` | Lag/time | NA-aware | active |
| — | Cyclic HMM (approx.) | `run_cyhmm()` | State-transition | NA-dropped | **disabled (slow)** |

Methods 3–4 (plus the disabled CyHMM) replace the original paper's Python CyHMM (see Overview). Lomb-Scargle and ACF are added because they tolerate the missing-day gaps in this dataset better than FFT.

**Time series structure:** For each participant-wave-signal, slots are ordered by day then by slot within day (slot_hour 0, 2, …, 22). Missing days leave gaps of 12 consecutive NAs. Each slot's time anchor in hours is `(day − 1) × 24 + slot_hour + 1` (midpoint of the 2-hr window).

```{r build-wave-ts}
build_wave_ts <- function(df_pw) {
  # df_pw: single participant-wave after imputation, sorted by day × slot_hour
  # Returns: ordered vector of signal values with NAs for missing-day gaps
  df_pw |>
    arrange(day, slot_hour) |>
    mutate(
      t_slot  = (day - 1L) * SLOTS_PER_DAY + (slot_hour / SLOT_HOURS) + 1L,
      t_hours = (day - 1L) * 24 + slot_hour + SLOT_HOURS / 2
    )
}
```

### 2.1 Fast Fourier Transform

FFT converts the signal to the frequency domain; the dominant frequency corresponds to the strongest periodic component (Section 3.1 of the paper). We detrend before applying FFT to remove slow drifts.

```{r fft-function}
run_fft <- function(x) {
  # x: contiguous time series vector (NAs linearly interpolated before FFT)
  x_filled <- approx(seq_along(x), x, seq_along(x))$y
  x_filled[is.na(x_filled)] <- mean(x_filled, na.rm = TRUE)
  x_detrend <- x_filled - predict(lm(x_filled ~ seq_along(x_filled)))

  n           <- length(x_detrend)
  fft_result  <- fft(x_detrend)
  power       <- Mod(fft_result)^2
  # frequency in cycles/slot
  freq_slot   <- (seq_len(n) - 1L) / n
  # period in hours = (1 / freq_slot) * SLOT_HOURS; exclude DC (idx=1) and mirror
  usable      <- 2:(n %/% 2)
  period_hrs  <- (1 / freq_slot[usable]) * SLOT_HOURS

  # Restrict to plausible biological periods (PERIOD_MIN_HR to PERIOD_MAX_HR)
  in_range    <- period_hrs >= PERIOD_MIN_HR & period_hrs <= PERIOD_MAX_HR
  if (!any(in_range)) return(NA_real_)

  dominant_idx   <- usable[in_range][which.max(power[usable][in_range])]
  dominant_period <- (1 / freq_slot[dominant_idx]) * SLOT_HOURS

  dominant_period
}
```

### 2.2 Chi-Squared Periodogram

Implements the Sokolove & Bushell Q_P statistic (Equation 1 in the paper). A period is considered significant when Q_P exceeds the chi-squared critical value with P−1 degrees of freedom at α = 0.01.

```{r chi-sq-function}
chi_sq_periodogram <- function(x) {
  # x: time series vector; NAs dropped (robust to irregular spacing)
  x_clean <- x[!is.na(x)]
  N       <- length(x_clean)
  M       <- mean(x_clean)
  denom   <- sum((x_clean - M)^2)
  if (denom == 0 || N < 2) return(NULL)

  periods_hr <- seq(PERIOD_MIN_HR, PERIOD_MAX_HR, by = PERIOD_STEP_HR)
  # Convert period in hours to period in slots (must be integer for segmenting)
  periods_sl <- round(periods_hr / SLOT_HOURS)

  qp_vals <- map_dbl(periods_sl, function(P_int) {
    K <- floor(N / P_int)
    if (K < 2) return(NA_real_)
    # Mh: mean of each within-period time unit across K full cycles
    x_mat <- matrix(x_clean[1:(K * P_int)], nrow = P_int, ncol = K)
    Mh    <- rowMeans(x_mat, na.rm = TRUE)
    (K * N * sum((Mh - M)^2)) / denom
  })

  # Significance: QP > chi-sq critical value with (P_int - 1) df
  sig_thresh <- qchisq(1 - CHI_ALPHA, df = pmax(periods_sl - 1L, 1L))
  is_sig     <- !is.na(qp_vals) & qp_vals > sig_thresh

  if (!any(is_sig)) {
    return(list(detected_period = NA_real_, qp_at_24hr = NA_real_, any_sig = FALSE))
  }

  # Return the significant period closest to 24 hours
  sig_periods <- periods_hr[is_sig]
  closest_idx <- which.min(abs(sig_periods - TARGET_PERIOD_HR))

  # Also record Q_P at the 24-hr period (rhythm strength)
  idx_24 <- which.min(abs(periods_hr - TARGET_PERIOD_HR))
  list(
    detected_period = sig_periods[closest_idx],
    qp_at_24hr      = qp_vals[idx_24],
    any_sig         = TRUE
  )
}
```

### 2.3 Lomb-Scargle Periodogram

The Lomb-Scargle periodogram (Lomb 1976; Scargle 1982) is a Fourier-based method designed for **unevenly-spaced data** — it operates directly on the observed (time, value) pairs without requiring imputation of missing days. This makes it the most robust of the active methods for our gappy data. We use the `lomb::lsp()` implementation and select the dominant peak within the biological range.

```{r lomb-function}
run_lomb <- function(x, t_hours) {
  # x: signal values; t_hours: time anchor in hours (same length, NAs dropped jointly)
  keep    <- !is.na(x) & !is.na(t_hours)
  x_obs   <- x[keep]
  t_obs   <- t_hours[keep]
  if (length(x_obs) < 10 || sd(x_obs) == 0) return(NA_real_)

  # Frequency range corresponds to PERIOD_MIN_HR..PERIOD_MAX_HR (cycles per hour)
  res <- tryCatch(
    lomb::lsp(
      x     = cbind(t_obs, x_obs),
      from  = 1 / PERIOD_MAX_HR,
      to    = 1 / PERIOD_MIN_HR,
      type  = "frequency",
      alpha = LS_ALPHA,
      plot  = FALSE
    ),
    error = function(e) NULL
  )
  if (is.null(res)) return(NA_real_)

  # Significant peak only (p-value below threshold), else NA
  if (res$p.value > LS_ALPHA) return(NA_real_)
  1 / res$peak.at[1]   # peak.at[1] is the frequency at the peak → period in hours
}
```

### 2.4 Autocorrelation (ACF)

Autocorrelation detects periodicity as the lag at which the signal correlates most strongly with itself. For a 24-hr rhythm at 2-hr resolution, we expect a peak at lag 12 slots. We search for the lag with maximum autocorrelation in the biological range and convert it to hours. `acf()` uses pairwise-complete handling for the NA gaps.

```{r acf-function}
run_acf <- function(x) {
  x_clean <- x
  if (sum(!is.na(x_clean)) < 2 * round(PERIOD_MAX_HR / SLOT_HOURS)) return(NA_real_)

  lag_min <- floor(PERIOD_MIN_HR / SLOT_HOURS)   # 6 slots  = 12 hr
  lag_max <- ceiling(PERIOD_MAX_HR / SLOT_HOURS) # 24 slots = 48 hr

  ac <- tryCatch(
    acf(x_clean, lag.max = lag_max, na.action = na.pass, plot = FALSE),
    error = function(e) NULL
  )
  if (is.null(ac)) return(NA_real_)

  # acf$acf[1] is lag 0; lags lag_min..lag_max are at indices (lag_min+1)..(lag_max+1)
  cand_lags <- lag_min:lag_max
  cand_vals <- ac$acf[cand_lags + 1L]

  # Require a positive local-correlation peak
  if (all(is.na(cand_vals)) || max(cand_vals, na.rm = TRUE) <= 0) return(NA_real_)

  best_lag <- cand_lags[which.max(cand_vals)]
  best_lag * SLOT_HOURS   # period in hours
}
```

### 2.5 Cyclic HMM Approximation (depmixS4)

A true CyHMM (Pierson et al. 2018) infers cyclic latent states whose count equals the period length, with Poisson dwell times and transitions constrained to a fixed cyclic order. The original Python implementation is Python-2 only and depends on a deprecated `pomegranate` API, so we approximate it with a standard Gaussian HMM (`depmixS4`): we fit models with a range of state counts (each candidate state count = candidate period in slots) and select the period whose model minimizes BIC. This captures the "number of recurring regimes per cycle" idea while staying in R.

> **⚠️ CyHMM is temporarily DISABLED** (`eval = FALSE`) to avoid the slow `depmixS4` EM step. The function and its call site below are kept (commented out) for easy re-enabling. Periodicity detection currently runs on the **four** other methods (FFT, Chi-squared, Lomb-Scargle, ACF).

```{r cyhmm-function, eval = FALSE}
run_cyhmm <- function(x, min_states = round(PERIOD_MIN_HR / SLOT_HOURS),
                          max_states = round(PERIOD_MAX_HR / SLOT_HOURS)) {
  # At 2-hr resolution: min_states = 6 (12 hr), max_states = 24 (48 hr)
  x_clean <- x[!is.na(x)]
  if (length(x_clean) < max_states * 2 || sd(x_clean) == 0) return(NA_real_)

  state_range <- seq(min_states, max_states)
  bic_vals <- map_dbl(state_range, function(ns) {
    tryCatch({
      mod <- depmix(
        response = list(x_clean ~ 1),
        data     = data.frame(x_clean = x_clean),
        nstates  = ns,
        family   = list(gaussian())
      )
      # depmixS4::fit() prints "converged at iteration ... with logLik:" via cat();
      # verbose=FALSE does NOT silence it, so capture & discard that console output.
      fitted_mod <- NULL
      invisible(utils::capture.output(
        fitted_mod <- fit(mod, verbose = FALSE, emcontrol = em.control(maxit = 200))
      ))
      suppressWarnings(BIC(fitted_mod))
    }, error = function(e) NA_real_)
  })

  if (all(is.na(bic_vals))) return(NA_real_)
  best_states <- state_range[which.min(bic_vals)]
  best_states * SLOT_HOURS   # return period in hours
}
```

### 2.6 Period Confirmation (≥3-of-4 consensus)

A 24-hr period is confirmed when **at least `CONSENSUS_FRAC` (75%) of the non-NA methods** report a period within ±`PERIOD_TOL_HR` hours of the 24-hr target. With all 4 active methods reporting, ≥3 must agree; if a method returns NA (could not estimate), it is excluded from both numerator and denominator. The confirmed period is the mean of the agreeing estimates.

```{r confirm-period}
confirm_period <- function(periods, target = TARGET_PERIOD_HR,
                           tol = PERIOD_TOL_HR, frac = CONSENSUS_FRAC) {
  # periods: named numeric vector of per-method period estimates (may contain NA)
  valid <- periods[!is.na(periods)]
  if (length(valid) < 2) return(NA_real_)

  near_target <- abs(valid - target) <= tol
  if (mean(near_target) >= frac) {
    return(mean(valid[near_target]))   # confirmed: mean of agreeing methods
  }
  NA_real_                              # no consensus
}
```

### Apply periodicity detection

```{r apply-periodicity}
# Nest data per participant × wave × signal, then apply the four active methods
ts_nested <- dat_imputed |>
  mutate(is_weekend = dt_wknd) |> 
  select(participant_id, wave, day, slot_hour, is_weekend, all_of(SIGNALS)) |>
  pivot_longer(all_of(SIGNALS), names_to = "signal", values_to = "value") |>
  group_by(participant_id, wave, signal) |>
  arrange(day, slot_hour, .by_group = TRUE) |>
  nest() |>
  ungroup()

# Pull out the per-(participant × wave × signal) time series and time axis once
ts_vec  <- map(ts_nested$data, \(d) d$value)
t_hours <- map(ts_nested$data, \(d) (d$day - 1L) * 24 + d$slot_hour + SLOT_HOURS / 2)

# Run the FOUR active methods, each with its own named progress bar.
# purrr's `.progress` shows a labelled bar per call.
# (Cyclic HMM is temporarily disabled — see the commented block below.)
fft_period   <- map_dbl(ts_vec, run_fft,            .progress = "1/4 FFT")
chi_result   <- map(ts_vec, chi_sq_periodogram,     .progress = "2/4 Chi-squared")
lomb_period  <- map2_dbl(ts_vec, t_hours, run_lomb, .progress = "3/4 Lomb-Scargle")
acf_period   <- map_dbl(ts_vec, run_acf,            .progress = "4/4 ACF")

# Derive chi-squared columns from its (cheap) list output
chi_period <- map_dbl(chi_result, \(r) if (is.null(r)) NA_real_ else r$detected_period)
qp_at_24hr <- map_dbl(chi_result, \(r) if (is.null(r)) NA_real_ else r$qp_at_24hr)

# ── CyHMM TEMPORARILY DISABLED ────────────────────────────────────────────────
# The lazy-CyHMM block below is commented out to skip the depmixS4 bottleneck.
# To re-enable: re-enable the run_cyhmm chunk above (remove eval = FALSE), uncomment
# this block, and add `cyhmm = cyhmm_period` back into `method_mat`.
#
# near_target <- function(p) !is.na(p) & abs(p - TARGET_PERIOD_HR) <= PERIOD_TOL_HR
# consensus_reached <- function(n_near, n_valid) {
#   n_valid >= 2 && (n_near / n_valid) >= CONSENSUS_FRAC
# }
# cheap_mat <- cbind(fft = fft_period, chi = chi_period,
#                    lomb = lomb_period, acf = acf_period)
# cyhmm_decisive <- map_lgl(seq_len(nrow(cheap_mat)), function(i) {
#   v   <- cheap_mat[i, ]
#   m_c <- sum(!is.na(v))          # cheap methods that reported
#   a_c <- sum(near_target(v))     # cheap methods near 24 h
#   b_na   <- consensus_reached(a_c,     m_c)       # CyHMM not run
#   b_near <- consensus_reached(a_c + 1, m_c + 1)   # CyHMM agrees
#   b_far  <- consensus_reached(a_c,     m_c + 1)   # CyHMM disagrees
#   !(b_na == b_near && b_na == b_far)              # decisive iff outcomes differ
# })
# message(sprintf(
#   "CyHMM decisive for %d of %d series (%.1f%%) — skipping the other %d.",
#   sum(cyhmm_decisive), length(cyhmm_decisive),
#   100 * mean(cyhmm_decisive), sum(!cyhmm_decisive)
# ))
# cyhmm_period <- rep(NA_real_, nrow(cheap_mat))
# cyhmm_period[cyhmm_decisive] <- map_dbl(
#   ts_vec[cyhmm_decisive], run_cyhmm, .progress = "5/5 Cyclic HMM (decisive only)"
# )

# Assemble per-method estimates (4 active methods) and apply the consensus rule
# (CONSENSUS_FRAC = 0.75 → ≥3 of 4 methods must agree).
method_mat <- cbind(fft = fft_period, chi = chi_period,
                    lomb = lomb_period, acf = acf_period)

periodicity_results <- ts_nested |>
  mutate(
    fft_period   = fft_period,
    chi_period   = chi_period,
    qp_at_24hr   = qp_at_24hr,
    lomb_period  = lomb_period,
    acf_period   = acf_period,
    # cyhmm_period = cyhmm_period,   # CyHMM temporarily disabled
    confirmed_period = apply(method_mat, 1, confirm_period),
    n_methods_reporting = apply(method_mat, 1, \(v) sum(!is.na(v))),
    n_methods_agree     = apply(method_mat, 1, \(v) {
      v <- v[!is.na(v)]; sum(abs(v - TARGET_PERIOD_HR) <= PERIOD_TOL_HR)
    }),
    is_circadian = !is.na(confirmed_period) &
                   abs(confirmed_period - TARGET_PERIOD_HR) <= PERIOD_TOL_HR
  ) |>
  select(-data) |>
  glimpse()
```

```{r periodicity-summary}
# Per-method detection rates + overall confirmation rate
detect_near <- function(p) round(mean(!is.na(p) & abs(p - TARGET_PERIOD_HR) <= PERIOD_TOL_HR) * 100, 1)

periodicity_results |>
  group_by(wave, signal) |>
  summarise(
    n             = n(),
    FFT           = detect_near(fft_period),
    `Chi-sq`      = detect_near(chi_period),
    `Lomb-Scargle` = detect_near(lomb_period),
    ACF           = detect_near(acf_period),
    # CyHMM       = detect_near(cyhmm_period),   # CyHMM temporarily disabled
    `Confirmed (≥3/4)` = round(mean(is_circadian) * 100, 1),
    .groups = "drop"
  ) |>
  ktab(
    "Periodicity detection results by method",
    subtitle = "per-method % of participant-waves detecting a 24-hr period (±4 hr); last column = ≥3-of-4 consensus"
  )
```

```{r consensus-distribution}
# How many methods agreed per participant-wave? Shows where consensus lands
periodicity_results |>
  count(wave, signal, n_methods_agree) |>
  ggplot(aes(x = factor(n_methods_agree), y = n, fill = n_methods_agree >= 3)) +
  geom_col() +
  facet_grid(wave ~ signal) +
  scale_fill_manual(values = c(`FALSE` = "grey70", `TRUE` = "steelblue"),
                    labels = c("< 3 (not confirmed)", "≥ 3 (confirmed)"),
                    name = NULL) +
  labs(
    x = "Number of methods detecting a 24-hr period (of 4)",
    y = "N participant-waves",
    title = "Method agreement per participant-wave",
    subtitle = "Blue = meets ≥3-of-4 consensus threshold"
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```

```{r periodicity-plot}
periodicity_results |>
  filter(!is.na(confirmed_period)) |>
  ggplot(aes(x = confirmed_period, fill = signal)) +
  geom_histogram(bins = 30, position = "identity", alpha = 0.6) +
  geom_vline(xintercept = 24, linetype = "dashed", colour = "black") +
  facet_grid(wave ~ signal) +
  labs(
    x = "Confirmed period (hours)",
    y = "N participant-waves",
    title = "Distribution of confirmed periods by wave and signal"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

---

<!-- ## Stage 3: Change-Point Detection -->

<!-- Following the paper's Stage 2 (Section 3.2), we identify change points (regime shifts) in the time series. At 2-hr resolution we apply the detector to the **full wave time series** (~252 time points) rather than per-day (which would have only 12 points — insufficient). Missing days are represented as NA gaps, which Rbeast handles natively. -->

<!-- **AutoNOM is disabled** (its minute-level frequency-change localization adds little at 2-hr resolution); **Rbeast is the sole change-point method**, tuned for speed (see §3.2). We restrict change-point analysis to participant-waves with a confirmed 24-hr period. -->

<!-- ### 3.1 AutoNOM -->

<!-- AutoNOM (Hadj-Amar et al., 2019) fits non-stationary sinusoidal segments via reversible-jump MCMC. `kmax` is selected using the MAPE elbow method (Table 4 in the paper). -->

<!-- > **⚠️ AutoNOM is DISABLED** (`eval = FALSE`). Its fine-grained frequency-change localization needs minute-level data; at 2-hr resolution it adds little over Rbeast and is slow. The function is kept (commented) for reference. **Rbeast (§3.2) is the sole change-point method.** -->

<!-- **Installation (optional):** `remotes::install_github("hadjamar/BayesSpec")`. -->

<!-- ```{r autonOM-function, eval = FALSE} -->
<!-- # BayesSpec (AutoNOM) is a GitHub-only package and is OPTIONAL: if it is not -->
<!-- # installed, AutoNOM is skipped and Rbeast (§3.2) carries the change-point analysis. -->
<!-- # The package name is held in a variable and resolved with getExportedValue() so -->
<!-- # that renv's static dependency scanner does NOT try to auto-install it from CRAN -->
<!-- # (where it does not exist). Install manually if you want AutoNOM: -->
<!-- #   remotes::install_github("hadjamar/BayesSpec") -->
<!-- BAYESSPEC_PKG  <- "BayesSpec" -->
<!-- BAYESSPEC_OK   <- requireNamespace(BAYESSPEC_PKG, quietly = TRUE) -->
<!-- autnom_fun     <- if (BAYESSPEC_OK) getExportedValue(BAYESSPEC_PKG, "autnom") else NULL -->

<!-- run_autonOM <- function(x, kmax_range = KMAX_RANGE, mmax = MMAX) { -->
<!--   empty <- list(changepoints = integer(0), fitted = rep(NA_real_, length(x)), -->
<!--                 kmax_used = NA_integer_, mape = NA_real_) -->
<!--   if (!BAYESSPEC_OK) return(empty)   # AutoNOM unavailable → Rbeast handles it -->

<!--   # Linearly interpolate short gaps (≤ 3 × SLOTS_PER_DAY missing slots) -->
<!--   n_consec_na <- rle(is.na(x)) -->
<!--   max_gap     <- max(n_consec_na$lengths[n_consec_na$values], default = 0L) -->

<!--   x_interp <- if (max_gap <= 3L * SLOTS_PER_DAY) { -->
<!--     approx(seq_along(x), x, seq_along(x))$y -->
<!--   } else { -->
<!--     x  # leave long gaps; AutoNOM may struggle — Rbeast preferred in this case -->
<!--   } -->
<!--   x_interp[is.na(x_interp)] <- mean(x_interp, na.rm = TRUE) -->

<!--   # MAPE-based kmax selection (Section 3 / Table 4 of paper) -->
<!--   mape_vals <- map_dbl(kmax_range, function(k) { -->
<!--     tryCatch({ -->
<!--       res <- autnom_fun(y = x_interp, kmax = k, mmax = mmax) -->
<!--       mean(abs((res$fitted - x_interp) / (x_interp + 1e-8))) * 100 -->
<!--     }, error = function(e) NA_real_) -->
<!--   }) -->

<!--   # Elbow: largest drop in MAPE as k increases -->
<!--   if (all(is.na(mape_vals))) return(empty) -->

<!--   diffs     <- diff(mape_vals) -->
<!--   best_k    <- kmax_range[which.min(diffs) + 1L] -->
<!--   best_res  <- autnom_fun(y = x_interp, kmax = best_k, mmax = mmax) -->

<!--   list( -->
<!--     changepoints = best_res$changepoints, -->
<!--     fitted       = best_res$fitted, -->
<!--     kmax_used    = best_k, -->
<!--     mape         = min(mape_vals, na.rm = TRUE) -->
<!--   ) -->
<!-- } -->
<!-- ``` -->

<!-- ### 3.2 Rbeast -->

<!-- Rbeast decomposes the time series into trend and seasonal components and estimates change-point probabilities. It handles irregular/missing data natively and is the sole change-point method here. The call is tuned for speed: MCMC draws are cut from the defaults (samples 8000→3000, burnin 200→150, chains 3→2) and the trend/season change-point counts are capped (`tcp.minmax = c(0,6)`, `scp.minmax = c(0,3)`) — we only need a disruption count, so a small model space samples much faster while leaving the 0.5-threshold change-point probabilities stable. -->

<!-- ```{r rbeast-function} -->
<!-- run_rbeast <- function(x) { -->
<!--   # period = 12 slots (24-hr circadian period at 2-hr resolution). NA gaps are -->
<!--   # handled automatically. Runtime is tuned down from beast()'s defaults: -->
<!--   #   • mcmc.samples 8000 → 3000, burnin 200 → 150, chains 3 → 2  (~4× fewer draws) -->
<!--   #   • tcp/scp.minmax cap the number of trend/season change points — we only need a -->
<!--   #     disruption count, so a small model space is enough and much faster to sample. -->
<!--   # cpPr (change-point probability) is robust at a 0.5 threshold under this budget. -->
<!--   tryCatch({ -->
<!--     result <- beast( -->
<!--       x, -->
<!--       season         = "harmonic", -->
<!--       period         = SLOTS_PER_DAY,   # 12 slots = 24 hours -->
<!--       tcp.minmax     = c(0, 6),         # ≤6 trend change points -->
<!--       scp.minmax     = c(0, 3),         # ≤3 seasonal change points -->
<!--       mcmc.samples   = 3000, -->
<!--       mcmc.burnin    = 150, -->
<!--       mcmc.chains    = 2, -->
<!--       print.progress = FALSE, -->
<!--       print.warning  = FALSE, -->
<!--       quiet          = TRUE -->
<!--     ) -->
<!--     # Change points where posterior probability > 0.5 -->
<!--     cp_idx <- which(result$trend$cpPr > 0.5) -->
<!--     list( -->
<!--       changepoints = cp_idx, -->
<!--       cp_prob      = result$trend$cpPr, -->
<!--       trend_fitted = result$trend$Y -->
<!--     ) -->
<!--   }, error = function(e) { -->
<!--     list(changepoints = integer(0), cp_prob = NA, trend_fitted = NA) -->
<!--   }) -->
<!-- } -->
<!-- ``` -->

<!-- ### Apply change-point detection -->

<!-- ```{r apply-changepoint} -->
<!-- # Restrict to confirmed 24-hr period participant-waves -->
<!-- pw_confirmed <- periodicity_results |> -->
<!--   filter(is_circadian) |> -->
<!--   select(participant_id, wave, signal) -->

<!-- cp_input <- ts_nested |> -->
<!--   semi_join(pw_confirmed, by = c("participant_id", "wave", "signal")) |> -->
<!--   mutate(ts_vec = map(data, \(d) d$value)) -->

<!-- # AutoNOM is disabled (see §3.1); Rbeast is the sole change-point method. -->
<!-- rbeast_res <- map( -->
<!--   cp_input$ts_vec,  -->
<!--   run_rbeast,  -->
<!--   .progress = "Rbeast change-points" -->
<!-- ) -->

<!-- cp_results <- cp_input |> -->
<!--   mutate( -->
<!--     cp_rbeast   = map(rbeast_res, \(r) r$changepoints), -->
<!--     n_cp_rbeast = map_int(cp_rbeast, length) -->
<!--   ) |> -->
<!--   select(-data, -ts_vec) -->
<!-- ``` -->

<!-- ```{r changepoint-summary} -->
<!-- cp_results |> -->
<!--   group_by(wave, signal) |> -->
<!--   summarise( -->
<!--     n                  = n(), -->
<!--     median_cp_rbeast   = median(n_cp_rbeast, na.rm = TRUE), -->
<!--     mean_cp_rbeast     = round(mean(n_cp_rbeast, na.rm = TRUE), 2), -->
<!--     pct_zero_cp_rbeast = round(mean(n_cp_rbeast == 0) * 100, 1), -->
<!--     .groups = "drop" -->
<!--   ) |> -->
<!--   ktab( -->
<!--     "Change-point detection summary (Rbeast)", -->
<!--     subtitle = "full wave series (~252 points); restricted to confirmed 24-hr participants" -->
<!--   ) -->
<!-- ``` -->

<!-- ```{r changepoint-plot} -->
<!-- cp_results |> -->
<!--   ggplot(aes(x = n_cp_rbeast)) + -->
<!--   geom_bar(fill = "steelblue") + -->
<!--   facet_grid(wave ~ signal) + -->
<!--   labs(x = "Number of change points (Rbeast)", y = "N participant-waves", -->
<!--        title = "Distribution of change-point counts per participant-wave") + -->
<!--   theme_minimal() -->
<!-- ``` -->

---

## Stage 4: Feature Extraction

For each participant × wave × signal with a confirmed 24-hr period, we extract:

1. **Cosinor parameters** — MESOR, amplitude, acrophase (via linearized least-squares)
2. **Rhythm strength** — Q_P at 24 hr from the chi-squared periodogram
3. **Non-parametric circadian features** — IS (interdaily stability), IV (intradaily variability), RA (relative amplitude)
4. **Change-point features** — number of Rbeast change points per wave, SD of change-point timing

```{r cosinor-function}
fit_cosinor <- function(x, t_hours) {
  # Linearized cosinor: y = a + b*cos(2π*t/24) + c*sin(2π*t/24)
  # MESOR = a, amplitude = sqrt(b²+c²), acrophase = atan2(-c, b) mapped to [0,24)
  df <- data.frame(
    y    = x,
    t    = t_hours,
    cosx = cos(2 * pi * t_hours / TARGET_PERIOD_HR),
    sinx = sin(2 * pi * t_hours / TARGET_PERIOD_HR)
  ) |> filter(!is.na(y))

  if (nrow(df) < 3) return(rep(NA_real_, 3) |> setNames(c("mesor", "amplitude", "acrophase")))

  fit <- tryCatch(
    lm(y ~ cosx + sinx, data = df),
    error = function(e) NULL
  )
  if (is.null(fit)) return(rep(NA_real_, 3) |> setNames(c("mesor", "amplitude", "acrophase")))

  b   <- coef(fit)[["cosx"]]
  cc  <- coef(fit)[["sinx"]]
  amp <- sqrt(b^2 + cc^2)
  aph <- (atan2(-cc, b) * 24 / (2 * pi)) %% 24  # convert to hours [0,24)

  c(mesor = unname(coef(fit)[["(Intercept)"]]), amplitude = amp, acrophase = aph)
}
```

```{r nonparam-circadian-functions}
# IS: interdaily stability — consistency of 24-hr pattern across days
# IV: intradaily variability — within-day fragmentation
# RA: relative amplitude = (M10 - L5) / (M10 + L5)

calc_is_iv <- function(x, n_slots_per_day = SLOTS_PER_DAY) {
  x_clean <- x[!is.na(x)]
  N <- length(x_clean)
  if (N < n_slots_per_day) return(c(IS = NA_real_, IV = NA_real_))

  M   <- mean(x_clean)
  var_total <- sum((x_clean - M)^2)
  if (var_total == 0) return(c(IS = NA_real_, IV = NA_real_))

  # IS: mean value at each within-period slot position
  slot_idx <- ((seq_along(x_clean) - 1L) %% n_slots_per_day) + 1L
  Mh  <- tapply(x_clean, slot_idx, mean)
  IS  <- N * sum((Mh - M)^2) / (n_slots_per_day * var_total)

  # IV: successive differences (only within valid stretches, not across NA gaps)
  diffs <- diff(x_clean)
  IV <- N * sum(diffs^2) / ((N - 1L) * var_total)

  c(IS = IS, IV = IV)
}

calc_ra <- function(x) {
  x_clean <- x[!is.na(x)]
  if (length(x_clean) < M10_SLOTS + L5_SLOTS) return(c(M10 = NA_real_, L5 = NA_real_, RA = NA_real_))

  # Rolling-window mean (slot-level, across the full wave).
  # Use stats::filter explicitly — dplyr::filter masks it on the search path.
  roll_sum <- function(v, w) stats::filter(v, rep(1 / w, w), sides = 1)

  m10_vals <- roll_sum(x_clean, M10_SLOTS)
  l5_vals  <- roll_sum(x_clean, L5_SLOTS)

  M10 <- max(m10_vals, na.rm = TRUE)
  L5  <- min(l5_vals,  na.rm = TRUE)
  RA  <- if (M10 + L5 == 0) NA_real_ else (M10 - L5) / (M10 + L5)

  c(M10 = M10, L5 = L5, RA = RA)
}
```

```{r extract-features}
# Build time axis vectors — full series and day-type sub-series — for cosinor fitting
ts_with_time <- ts_nested |>
  # semi_join(pw_confirmed, by = c("participant_id", "wave", "signal")) |>
  mutate(
    # Full series
    t_hours    = map(data, \(d) (d$day - 1L) * 24 + d$slot_hour + SLOT_HOURS / 2),
    ts_vec     = map(data, \(d) d$value),
    # Weekday slots (Mon–Fri)
    t_hours_wd = map(data, \(d) {
      idx <- !d$is_weekend
      (d$day[idx] - 1L) * 24 + d$slot_hour[idx] + SLOT_HOURS / 2
    }),
    ts_vec_wd  = map(data, \(d) d$value[!d$is_weekend]),
    # Weekend slots (Sat–Sun)
    t_hours_we = map(data, \(d) {
      idx <- d$is_weekend
      (d$day[idx] - 1L) * 24 + d$slot_hour[idx] + SLOT_HOURS / 2
    }),
    ts_vec_we  = map(data, \(d) d$value[d$is_weekend])
  )

cosinor_features <- ts_with_time |>
  mutate(
    # Full-series cosinor (amplitude/mesor over full wave)
    cos_full      = map2(ts_vec,    t_hours,    fit_cosinor),
    mesor         = map_dbl(cos_full, "mesor"),
    amplitude     = map_dbl(cos_full, "amplitude"),
    acrophase     = map_dbl(cos_full, "acrophase"),
    # Weekday cosinor
    cos_wd         = map2(ts_vec_wd, t_hours_wd, fit_cosinor),
    mesor_wday     = map_dbl(cos_wd, "mesor"),
    amplitude_wday = map_dbl(cos_wd, "amplitude"),
    acrophase_wday = map_dbl(cos_wd, "acrophase"),
    # Weekend cosinor
    cos_we         = map2(ts_vec_we, t_hours_we, fit_cosinor),
    mesor_wend     = map_dbl(cos_we, "mesor"),
    amplitude_wend = map_dbl(cos_we, "amplitude"),
    acrophase_wend = map_dbl(cos_we, "acrophase"),
    # Social jetlag: signed circular difference wrapped to [-12, 12] hours
    # Positive = weekend peak is later than weekday peak (typical adolescent pattern)
    social_jetlag  = ((acrophase_wend - acrophase_wday + 12) %% 24) - 12
  ) |>
  select(participant_id, wave, signal,
         mesor, amplitude, acrophase,
         mesor_wday, amplitude_wday, acrophase_wday,
         mesor_wend, amplitude_wend, acrophase_wend,
         social_jetlag)

nonparam_features <- ts_with_time |>
  mutate(
    is_iv = map(ts_vec, calc_is_iv),
    ra    = map(ts_vec, calc_ra),
    IS    = map_dbl(is_iv, "IS"),
    IV    = map_dbl(is_iv, "IV"),
    M10   = map_dbl(ra, "M10"),
    L5    = map_dbl(ra, "L5"),
    RA    = map_dbl(ra, "RA")
  ) |>
  select(participant_id, wave, signal, IS, IV, M10, L5, RA)

# cp_features <- cp_results |>
#   mutate(
#     cp_timing_sd_rbeast = map_dbl(cp_rbeast, \(cp) if (length(cp) < 2) NA_real_ else sd(cp))
#   ) |>
#   select(participant_id, wave, signal, n_cp_rbeast, cp_timing_sd_rbeast)

# Rhythm strength from chi-sq
rhythm_strength <- periodicity_results |>
  filter(is_circadian) |>
  select(participant_id, wave, signal, qp_at_24hr)

# Master feature table (one row per participant × wave × signal)
features <- cosinor_features |>
  left_join(nonparam_features, by = c("participant_id", "wave", "signal")) |>
  # left_join(cp_features,       by = c("participant_id", "wave", "signal")) |>
  left_join(rhythm_strength,   by = c("participant_id", "wave", "signal")) |>
  left_join(
    periodicity_results |> select(participant_id, wave, signal, confirmed_period, is_circadian),
    by = c("participant_id", "wave", "signal")
  ) |>
  mutate(arrhythmic = !is_circadian) |> 
  glimpse()
```

```{r feature-summary}
features |>
  group_by(wave, signal) |>
  summarise(
    n              = sum(!arrhythmic),
    pct_arrhythmic = round(mean(arrhythmic) * 100, 1),
    across(
      c(amplitude_wday, amplitude_wend, acrophase_wday, acrophase_wend,
        social_jetlag, IS, IV, RA, qp_at_24hr),
      \(x) round(median(x, na.rm = TRUE), 2),
      .names = "med_{.col}"
    ),
    .groups = "drop"
  ) |>
  ktab(
    "Feature summary by wave and signal (medians)",
    subtitle = "Cosinor weekday/weekend · Social jetlag · Non-parametric (IS/IV/RA) · Rhythm strength (qp)"
  )
```

---

## Stage 5: Longitudinal Modeling

Growth curve models (Bates et al., 2015; `lme4`) test whether rhythm features change across adolescence and whether trajectories differ by sex. The approximate ages at each wave are 12 (yr2), 14 (yr4), 16 (yr6). Participant random intercepts and slopes account for individual differences.

```{r longitudinal-setup}
# Join approximate age; add sin/cos projections for circular acrophase
features_long <- features |>
  mutate(
    age      = WAVE_AGES[wave],
    age_c    = age - 12,        # centered at youngest wave (yr2 ≈ 12)
    age_c_sq = age_c^2,
    # Circular projections for acrophase — avoids 0/24 h boundary artefacts in lmer
    sin_aph_wday = sin(2 * pi * acrophase_wday / 24),
    cos_aph_wday = cos(2 * pi * acrophase_wday / 24),
    sin_aph_wend = sin(2 * pi * acrophase_wend / 24),
    cos_aph_wend = cos(2 * pi * acrophase_wend / 24)
  )

# Z-score continuous features across all participant-waves (pooled)
# Acrophase excluded — use its sin/cos projections above instead
z_cols <- c("mesor_wday", "mesor_wend",
            "amplitude_wday", "amplitude_wend",
            "social_jetlag", "IS", "IV", "RA", "qp_at_24hr")

features_long <- features_long |>
  mutate(across(all_of(z_cols),
                \(x) as.numeric(scale(x)),
                .names = "{.col}_z"))

# Uncomment and adapt if a demographics file with sex/puberty is available:
# demo <- read_csv("data/vars-of-interest/demographics.csv")
# features_long <- features_long |> left_join(demo, by = "participant_id")
```

### 5.1 Circadian phase (acrophase)

Acrophase is a circular variable (0–24 h). Fitting a linear model directly is biased near the 0/24 h wrap. Instead, we project acrophase onto its unit-circle components — sin(2π·aph/24) and cos(2π·aph/24) — and fit each with a separate LMM. The direction of developmental change (phase delay vs advance) is recovered from the sign of the mean vector across waves.

```{r acrophase-plot}
features_long |>
  filter(!is.na(acrophase_wday), !is.na(acrophase_wend), !arrhythmic) |>
  glimpse()
  pivot_longer(c(acrophase_wday, acrophase_wend),
               names_to = "day_type", values_to = "acrophase",
               names_prefix = "acrophase_") |>
  mutate(day_type = if_else(day_type == "wday", "Weekday", "Weekend")) |>
  ggplot(aes(x = acrophase, fill = wave)) +
  geom_density(alpha = 0.5) +
  facet_grid(day_type ~ signal) +
  scale_x_continuous(breaks = seq(0, 24, 4), limits = c(0, 24)) +
  geom_vline(xintercept = c(6, 18), linetype = "dotted") +
  labs(x = "Acrophase (hour of peak, 24-hr clock)", y = "Density",
       title = "Distribution of acrophase by wave, signal, and day type") +
  theme_bw(14)
```

```{r model-acrophase}
# Fit sin and cos components separately; random intercept only (sparse repeated measures)
fit_circ_lmer <- function(df_grp, outcome_col, signal_name) {
  tryCatch({
    f <- as.formula(paste(outcome_col, "~ age_c + (1 | participant_id)"))
    m <- lmer(f, data = df_grp, REML = FALSE,
              control = lmerControl(optimizer = "bobyqa"))
    broom.mixed::tidy(m, effects = "fixed") |>
      mutate(signal = signal_name, component = outcome_col)
  }, error = function(e) { warning(e$message); NULL })
}

aph_models <- features_long |>
  filter(!is.na(sin_aph_wday), !arrhythmic) |>
  group_by(signal) |>
  group_map(function(df, key) {
    bind_rows(
      fit_circ_lmer(df, "sin_aph_wday", key$signal),
      fit_circ_lmer(df, "cos_aph_wday", key$signal),
      fit_circ_lmer(df, "sin_aph_wend", key$signal),
      fit_circ_lmer(df, "cos_aph_wend", key$signal)
    )
  }) |>
  bind_rows()

aph_models |>
  filter(term == "age_c") |>
  select(signal, component, estimate, std.error, statistic, p.value) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  ktab("Acrophase components ~ age (fixed effect of age_c only)",
       subtitle = "sin/cos projections; positive sin slope = later peak with age; (1|participant_id)", digits = 3)
```

### 5.2 Social jetlag

Social jetlag (SJL) is the signed difference between weekend and weekday acrophase (wrapped to [−12, 12] h). A positive SJL indicates that the weekend activity peak is later than the weekday peak — the hallmark of circadian misalignment in adolescents. We test whether SJL changes across adolescence using the z-scored SJL feature.

```{r social-jetlag-plot}
features_long |>
  filter(!is.na(social_jetlag_z), !arrhythmic) |>
  ggplot(aes(x = social_jetlag_z, fill = wave)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~signal) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(x = "Social jetlag (h): weekend − weekday acrophase",
       y = "Density",
       title = "Distribution of social jetlag by wave and signal") +
  theme_bw(20)
```

```{r model-sjl}
features_long |>
  filter(!is.na(social_jetlag_z), !arrhythmic) |>
  group_by(signal) |>
  group_map(function(df, key) {
    tryCatch({
      m <- lmer(social_jetlag_z ~ age_c + (1 | participant_id), data = df,
                REML = FALSE, control = lmerControl(optimizer = "bobyqa"))
      broom.mixed::tidy(m, effects = "fixed") |> mutate(signal = key$signal)
    }, error = function(e) { warning(e$message); NULL })
  }) |>
  bind_rows() |>
  filter(term != "(Intercept)") |>
  select(signal, term, estimate, std.error, statistic, p.value) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  ktab("Social jetlag (z) ~ age (fixed effects)",
       subtitle = "Positive slope = increasing weekend–weekday misalignment with age; (1|participant_id)", digits = 3)
```

### 5.3 Rhythm amplitude

Does the peak-to-trough contrast in activity change across adolescence? Weekday and weekend amplitudes are modeled separately using z-scored values.

```{r amplitude-plot}
features |>
  filter(!is.na(amplitude_wday), !is.na(amplitude_wend), !arrhythmic) |>
  pivot_longer(c(amplitude_wday, amplitude_wend),
               names_to = "day_type", values_to = "amplitude",
               names_prefix = "amplitude_") |>
  mutate(day_type = if_else(day_type == "wday", "Weekday", "Weekend")) |>
  ggplot(aes(x = log10(amplitude + 1), fill = wave)) +
  geom_density(alpha = 0.5) +
  facet_grid(day_type ~ signal) +
  labs(x = "log10(Amplitude + 1)", y = "Density",
       title = "Distribution of amplitude by wave, signal, and day type") +
  theme_bw(14)
```

```{r model-amplitude}
bind_rows(
  features_long |>
    filter(!is.na(amplitude_wday_z), !arrhythmic) |>
    group_by(signal) |>
    group_map(function(df, key) {
      tryCatch({
        m <- lmer(amplitude_wday_z ~ age_c + (1 | participant_id), data = df,
                  REML = FALSE, control = lmerControl(optimizer = "bobyqa"))
        broom.mixed::tidy(m, effects = "fixed") |>
          mutate(signal = key$signal, outcome = "amplitude_wday")
      }, error = function(e) NULL)
    }) |> bind_rows(),
  features_long |>
    filter(!is.na(amplitude_wend_z), !arrhythmic) |>
    group_by(signal) |>
    group_map(function(df, key) {
      tryCatch({
        m <- lmer(amplitude_wend_z ~ age_c + (1 | participant_id), data = df,
                  REML = FALSE, control = lmerControl(optimizer = "bobyqa"))
        broom.mixed::tidy(m, effects = "fixed") |>
          mutate(signal = key$signal, outcome = "amplitude_wend")
      }, error = function(e) NULL)
    }) |> bind_rows()
) |>
  filter(term != "(Intercept)") |>
  select(outcome, signal, term, estimate, std.error, statistic, p.value) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  ktab("Amplitude (z) ~ age — weekday and weekend (fixed effects)", digits = 3)
```

### 5.4 Rhythm strength (Q_P)

Does the regularity/strength of the circadian rhythm change with age?

```{r model-strength}
features_long |>
  filter(!is.na(qp_at_24hr_z), !arrhythmic) |>
  group_by(signal) |>
  group_map(function(df, key) {
    tryCatch({
      m <- lmer(qp_at_24hr_z ~ age_c + (1 | participant_id), data = df,
                REML = FALSE, control = lmerControl(optimizer = "bobyqa"))
      broom.mixed::tidy(m, effects = "fixed") |> mutate(signal = key$signal)
    }, error = function(e) NULL)
  }) |>
  bind_rows() |>
  filter(term != "(Intercept)") |>
  select(signal, term, estimate, std.error, statistic, p.value) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  ktab("Rhythm strength QP (z) ~ age (fixed effects)", digits = 3)
```

### 5.5 Interdaily stability and intradaily variability

```{r model-is-iv}
bind_rows(
  features_long |> filter(!is.na(IS_z)) |>
    group_by(signal) |>
    group_map(function(df, key) {
      tryCatch({
        m <- lmer(IS_z ~ age_c + (1 | participant_id), data = df,
                  REML = FALSE, control = lmerControl(optimizer = "bobyqa"))
        broom.mixed::tidy(m, effects = "fixed") |>
          mutate(signal = key$signal, outcome = "IS")
      }, error = function(e) NULL)
    }) |> bind_rows(),
  features_long |> filter(!is.na(IV_z)) |>
    group_by(signal) |>
    group_map(function(df, key) {
      tryCatch({
        m <- lmer(IV_z ~ age_c + (1 | participant_id), data = df,
                  REML = FALSE, control = lmerControl(optimizer = "bobyqa"))
        broom.mixed::tidy(m, effects = "fixed") |>
          mutate(signal = key$signal, outcome = "IV")
      }, error = function(e) NULL)
    }) |> bind_rows()
) |>
  filter(term != "(Intercept)") |>
  select(outcome, signal, term, estimate, std.error, statistic, p.value) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  ktab("IS and IV (z) ~ age (fixed effects)", digits = 3)
```


## Session info

```{r session-info}
sessionInfo()
```
